---
title: "thesis_testing"
author: "Elias Ovesen"
format: html
editor: visual
---

I have run the same model as in the project thesis, but now with observation-level latent variable, and extended to 5 traits and 2 latent variables. This resulted in a poor run of the model, and horrible mixing. This is due to identifiability issues such as free rotation. The output is below.

\begin{table}[ht]
\centering
\caption{Inference Results for Stan Model (20,000 post-warmup draws)}
\label{tab:stan_results}
\small % Reduces font size slightly to fit all columns comfortably
\begin{tabular}{l r r r r r r r r}
\toprule
Parameter & mean & se\_mean & sd & 2.5\% & 50\% & 97.5\% & n\_eff & Rhat \\
\midrule
$\beta_{1,1}$ & -0.37 & 0.02 & 0.04 & -0.44 & -0.37 & -0.29 & 6 & 1.23 \\
$\beta_{1,2}$ & 0.10 & 0.01 & 0.05 & 0.01 & 0.10 & 0.19 & 38 & 1.04 \\
$\beta_{1,3}$ & 0.00 & 0.00 & 0.06 & -0.11 & 0.00 & 0.11 & 12985 & 1.00 \\
$\beta_{1,4}$ & 0.05 & 0.01 & 0.05 & -0.05 & 0.05 & 0.14 & 83 & 1.02 \\
$\beta_{1,5}$ & 0.06 & 0.00 & 0.04 & -0.01 & 0.05 & 0.13 & 110 & 1.02 \\
$\beta_{2,1}$ & 0.83 & 0.01 & 0.03 & 0.77 & 0.83 & 0.90 & 41 & 1.03 \\
$\beta_{2,2}$ & -0.08 & 0.00 & 0.04 & -0.16 & -0.08 & 0.00 & 72 & 1.03 \\
$\beta_{2,3}$ & 0.04 & 0.00 & 0.04 & -0.04 & 0.04 & 0.13 & 13208 & 1.00 \\
$\beta_{2,4}$ & 0.04 & 0.00 & 0.04 & -0.05 & 0.04 & 0.12 & 145 & 1.02 \\
$\beta_{2,5}$ & -0.03 & 0.00 & 0.04 & -0.10 & -0.03 & 0.04 & 2241 & 1.01 \\
\midrule
$\lambda_{1,1}$ & 0.41 & 0.17 & 0.23 & 0.00 & 0.53 & 0.58 & 2 & 17.11 \\
$\lambda_{1,2}$ & 0.04 & 0.14 & 0.20 & -0.23 & 0.06 & 0.27 & 2 & 12.83 \\
$\lambda_{2,1}$ & 0.39 & 0.46 & 0.66 & -0.78 & 0.75 & 0.81 & 2 & 35.78 \\
$\lambda_{2,2}$ & 0.11 & 0.07 & 0.10 & 0.00 & 0.08 & 0.25 & 2 & 6.36 \\
$\lambda_{3,1}$ & 0.00 & 0.02 & 0.03 & -0.05 & 0.00 & 0.05 & 2 & 2.19 \\
$\lambda_{3,2}$ & 0.00 & 0.68 & 0.96 & -1.00 & 0.00 & 0.99 & 2 & 50.90 \\
$\lambda_{4,1}$ & 0.41 & 0.53 & 0.75 & -0.92 & 0.83 & 0.88 & 2 & 38.67 \\
$\lambda_{4,2}$ & 0.09 & 0.05 & 0.07 & 0.00 & 0.08 & 0.20 & 2 & 3.61 \\
$\lambda_{5,1}$ & 0.23 & 0.27 & 0.38 & -0.46 & 0.44 & 0.49 & 2 & 22.36 \\
$\lambda_{5,2}$ & 0.05 & 0.21 & 0.30 & -0.28 & 0.04 & 0.38 & 2 & 18.39 \\
\midrule
var\_LV\_1\_gen & 0.32 & 0.01 & 0.04 & 0.24 & 0.32 & 0.39 & 36 & 1.05 \\
var\_LV\_1\_pe  & 0.38 & 0.02 & 0.04 & 0.30 & 0.38 & 0.46 & 7 & 1.20 \\
var\_LV\_1\_te  & 0.31 & 0.01 & 0.02 & 0.27 & 0.31 & 0.35 & 7 & 1.18 \\
var\_LV\_2\_gen & 0.49 & 0.00 & 0.05 & 0.39 & 0.49 & 0.57 & 2343 & 1.01 \\
var\_LV\_2\_pe  & 0.51 & 0.00 & 0.05 & 0.42 & 0.51 & 0.60 & 2381 & 1.01 \\
var\_LV\_2\_te  & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.01 & 680 & 1.02 \\
\midrule
$sd\_R[1]$ & 0.76 & 0.05 & 0.07 & 0.70 & 0.72 & 0.89 & 2 & 8.02 \\
$sd\_R[2]$ & 0.69 & 0.01 & 0.02 & 0.66 & 0.69 & 0.73 & 3 & 1.55 \\
$sd\_R[3]$ & 0.29 & 0.00 & 0.01 & 0.27 & 0.29 & 0.30 & 710 & 1.02 \\
$sd\_R[4]$ & 0.60 & 0.02 & 0.03 & 0.54 & 0.61 & 0.64 & 3 & 1.94 \\
$sd\_R[5]$ & 0.85 & 0.00 & 0.01 & 0.83 & 0.85 & 0.87 & 15 & 1.07 \\
\midrule
$h^2\_traits[1]$ & 0.11 & 0.03 & 0.05 & 0.02 & 0.13 & 0.17 & 2 & 3.96 \\
$h^2\_traits[3]$ & 0.45 & 0.00 & 0.04 & 0.36 & 0.45 & 0.53 & 2372 & 1.01 \\
$lp\_\_$ & -7347.03 & 236.28 & 362.03 & -8111.53 & -7229.68 & -6855.13 & 2 & 2.53 \\
\bottomrule
\end{tabular}
\end{table}

In the improved model I will not only handle these issues, but also implement a highly optimized version of the relationship structure, which uses a directed acyclical graph instead of the full A matrix. This uses a recursive method for computing the genetic random effects instead of evaluation the full A matrix (or its inverse) at every iteration. This should have been done earlier, as it likely drastically improves the runtime of the MCMC process. I also considered running a model (inspired by Paul Acker) which optimizes it even further by handling offspring (with no offspring) differently from the other individuals which do have offspring, as for the offspring it is more difficult to separate the genetic from the environmental effects due to having less genetic inference from them. But it seems that this combination is not as necessary for Stan and HMC as it is for MCMCglmm and Gibbs sampling. The former is better at navigating correlated parameters (which the genetic and environmental will be for the offspring) than the latter. In any case, it is likely that this optimization only improves the speed by a constant factor. However, the optimization of the recursive method (instead of the full A matrix) is absolutely necessary as it likely reduces the model evaluations from $O(n^3)$ to $O(n)$.

For two latent variables we have the additional identifiability issue of rotational freedom. If you have two loading vectors forming a plane, there is an infinite number of ways to rotate these within the plane and get the same data points, just changing the scalars multiplied with the rotated vectors (the latent variable scores). What happened when trying to run the model without any constraints except two positive components, is what can be seen in the data below. This is why it is not enough to just constrain one element of each vector to be positive; one of the components must be constrained to zero, in order to pin the "coordinate system" down and prevent rotation. The way this is done is by setting the loading of the second factor for the first trait to be zero, thus meaning that the first factor captures all the variation of the first trait (and the first trait is not captured at all by the second factor). This achieves an upper echelon form of the two factors.

However, when trying to constrain the first element of the second factor, which at the time corresponded to the wing length, to zero, meaning that the first component of the first factor and thus that only the first factor captures the wing length, the model still did not converge and we get lp values of neff = 2 and Rhat = 2.29. The results of this are below. This lack of convergence tells us that the wing length is not ideal, which could be because it does not have a strong enough genetic link (?). The heritability for the wing length is only 0.09 as estimated by this model, and the 95% credible interval for the first factor ($\lambda_{1,1}$) goes all the way down to zero, meaning it can be zero (elaborate/investigate more). The tarsus length (trait 3) seems to have a stronger genetic link with a heritability of 0.46 and it converged properly, so it might be useful to try using that as trait 1 (and loading on factor 1) instead of the wing length, so this will be attempted next.

\begin{table}[ht]
\centering
\small
\caption{Posterior Summaries: Documentation of Non-Converged Run. This is with the first component of the second factor set to zero, here corresponding to the wing length.}
\begin{tabular}{l S[table-format=-1.2] S[table-format=1.2] S[table-format=-1.2] S[table-format=-1.2] r S[table-format=2.2]}
\toprule
Parameter & {Mean} & {SD} & {2.5\%} & {97.5\%} & {$n_{eff}$} & {$\hat{R}$} \\
\midrule
$\beta[1,1]$ & -0.38 & 0.04 & -0.44 & -0.30 & 6 & 1.22 \\
$\beta[1,2]$ & 0.09 & 0.04 & 0.00 & 0.17 & 6441 & 1.00 \\
$\beta[1,3]$ & -0.01 & 0.06 & -0.12 & 0.10 & 12647 & 1.00 \\
$\beta[1,4]$ & 0.03 & 0.05 & -0.06 & 0.13 & 6796 & 1.00 \\
$\beta[1,5]$ & 0.05 & 0.03 & -0.02 & 0.11 & 8264 & 1.00 \\
$\beta[2,1]$ & 0.84 & 0.03 & 0.77 & 0.91 & 46 & 1.03 \\
$\beta[2,2]$ & -0.07 & 0.04 & -0.15 & 0.01 & 7923 & 1.00 \\
$\beta[2,3]$ & 0.05 & 0.04 & -0.04 & 0.13 & 12072 & 1.00 \\
$\beta[2,4]$ & 0.04 & 0.04 & -0.04 & 0.13 & 8223 & 1.00 \\
$\beta[2,5]$ & -0.03 & 0.03 & -0.09 & 0.04 & 11484 & 1.00 \\
\midrule
$\lambda[1,1]$ & 0.44 & 0.25 & 0.00 & 0.62 & 2 & 17.99 \\
$\lambda[1,2]$ & 0.00 & 0.00 & 0.00 & 0.00 & {---} & {---} \\
$\lambda[2,1]$ & 0.40 & 0.66 & -0.77 & 0.82 & 2 & 36.06 \\
$\lambda[2,2]$ & 0.03 & 0.01 & 0.01 & 0.06 & 4 & 1.46 \\
$\lambda[3,1]$ & 0.03 & 0.04 & -0.06 & 0.08 & 2 & 3.39 \\
$\lambda[3,2]$ & -0.48 & 0.83 & -0.99 & 0.98 & 2 & 43.37 \\
$\lambda[4,1]$ & 0.42 & 0.75 & -0.92 & 0.89 & 2 & 38.96 \\
$\lambda[4,2]$ & 0.04 & 0.06 & -0.07 & 0.12 & 2 & 4.30 \\
$\lambda[5,1]$ & 0.25 & 0.40 & -0.47 & 0.51 & 2 & 23.35 \\
$\lambda[5,2]$ & -0.11 & 0.20 & -0.27 & 0.25 & 2 & 13.09 \\
\midrule
$var\_LV\_1\_gen$ & 0.30 & 0.04 & 0.23 & 0.37 & 1313 & 1.02 \\
$var\_LV\_1\_pe$ & 0.41 & 0.04 & 0.34 & 0.49 & 29 & 1.06 \\
$var\_LV\_1\_temp$ & 0.29 & 0.02 & 0.25 & 0.33 & 30 & 1.05 \\
$var\_LV\_2\_gen$ & 0.50 & 0.05 & 0.40 & 0.58 & 1925 & 1.00 \\
$var\_LV\_2\_pe$ & 0.50 & 0.05 & 0.41 & 0.59 & 1787 & 1.01 \\
$var\_LV\_2\_temp$ & 0.01 & 0.01 & 0.00 & 0.02 & 476 & 1.02 \\
\midrule
$sd\_R[1]$ & 0.77 & 0.08 & 0.70 & 0.92 & 2 & 9.00 \\
$sd\_R[2]$ & 0.69 & 0.02 & 0.67 & 0.73 & 3 & 1.53 \\
$sd\_R[3]$ & 0.28 & 0.01 & 0.26 & 0.30 & 555 & 1.02 \\
$sd\_R[4]$ & 0.61 & 0.03 & 0.54 & 0.64 & 2 & 2.21 \\
$sd\_R[5]$ & 0.85 & 0.01 & 0.83 & 0.87 & 12 & 1.09 \\
\midrule
$h^2\_traits[1]$ & 0.09 & 0.05 & 0.00 & 0.15 & 2 & 4.18 \\
$h^2\_traits[2]$ & 0.17 & 0.02 & 0.13 & 0.21 & 2839 & 1.01 \\
$h^2\_traits[3]$ & 0.46 & 0.04 & 0.37 & 0.54 & 1878 & 1.00 \\
$h^2\_traits[4]$ & 0.20 & 0.03 & 0.15 & 0.26 & 11 & 1.12 \\
$h^2\_traits[5]$ & 0.09 & 0.01 & 0.07 & 0.11 & 2914 & 1.01 \\
\bottomrule
\end{tabular}
\end{table}

Now, with tarsus length as trait 1, I still got awful convergence on everything except the first loading on the first factor (for tarsus length). So this shows that tarsus length works, but using nebb h as the second trait might have not been a great idea, as the second factor is supposed to capture variance not captured by the first factor, and tarsus and beak may have too much in common (and the second trait defines the "anchor" of the second factor?). Therefore, I will try the body mass as the second trait in the next run. I also will tweak some hyperparameters (set max tree dept to 15 (instead of the standard 10) and also set the adapt delta parameter to 0.99). And also change the priors on the prop var (the simplex) from a uniform prior to a dirichlet prior with paramter 1.5, to slightly enforce the variance components away from zero. I also lowered the variance from 1 to 0.5 for the lambda priors (normal) and made the lambda 1,1 and lambda 2,2 (which are set positive) half-normal instead of lognormal to not push them away from zero (?).

Update: This did not work either, it in fact worked even worse than the beak height did. It can mean that the body mass is just explained by size (tarsus) and noise. For a final run without major changes I will run the model with tarsus as first and beak length as second trait, as the latter might be more of a shape trait than size, and then I increased the priors back to standard deviation 1 for the lambda priors.

If this works (it probably won't, though), the next step will likely be to extend the model to include survival (binary) for a number of years, number of offspring (Poisson likely) for the same number of years and also make some of the variables or responses time-varying, so this will be a significant extension of the model.

If it doesn't work, I will have to look into more ways to make it identifiable or consider whether 2 latent variables on these traits actually works.

Update: I got a tip from Bob: Use a different prior where the additive genetic variance of the second factor is based on that of the first factor, via a beta prior. This might make it work, even without forcing a certain component of the vector to be zero. Essentially, \begin{align*}
    \sigma_1^2 &\sim \text{Ga}(\dots) \\
    \sigma_2^2 &= \sigma_1^2 \cdot \rho, \\
    \rho &\sim \text{Beta}(1,1),
\end{align*} at least I think the parameters of the Beta distribution where 1, 1. This enforces an ordering of the sizes of the variance explained by the different latent variables. I might also be able to avoid enforcing a positive component in each factor to avoid sign flipping (this could also not be sufficient for reasons Bob explained on the whiteboard). In any case it is possible to handle the sign flipping after the estimations are already done, by mapping.

I also figured out what to do about the differently distributed traits. I kind of forgot how a GLM works, but you can just use different link functions for the different traits, but there is only one linear predictor (based on the latent variables). Essentially you make one "row" for each trait, each with its own link function linking the relevant data to the linear predictor. And then you can have multiple different link functions in one big model.

When it comes to identifiability, the best way to look at it I think is this: We only have the observed covariance matrix as information, which for 5 traits gives us 15 unique values as it's symmetric. Then the values we need to estimate is according to the factor analytic equation \eqref{cov_mat_rank_red}, in which you have a loading term and a residual term. The assumption in the model is that all covariance is accounted for by the first term. This means that the residual matrix is assumed to be diagonal. This was the issue with the individual-level latent variable, because it didn't account for any shared variance between the trait that was only tied to the measurement level, not permanently to the individual. But I think this, that the residual matrix is diagonal, is a central assumption of the factor analytic framework. In any case, the information consists of 15 unique values, when combining all the information from all the data points for all 5 traits into the "statistics" from them. Then, the parameters to estimate are the loadings and the elements of the residual matrix in \eqref{cov_mat_rank_red}. The number of loadings equals the number of traits times the number of latent variables, so with two LVs this is 10, and then 5 residuals. This is also 15 unique values, making the model barely identifiable in this case, or with zero degrees of freedom. If we constrain one loading to be zero, for example, to prevent rotational freedom, this removes one parameter to estimate, so now it is 14. This either adds or removes a degree of freedom (not sure exactly how this works). This explains why, for example, fitting 3 LV's on these 5 traits won't work. When I had 2 traits and 1 LV, I actually had only 3 elements of information but had to estimate 2 + 2 variables, so it shouldn't really have been identified. But maybe the priors or data helped it converge, but I guess maybe that this was not the unique solution in that case. This could be investigated further later on. When it comes to the other values we are estimating, or sampling, like the scores, these are just random effects, and are sampled from a standard normal distribution, so these don't count as parameters to estimate. So I also should think about the relationship between the use of priors for these, for the variance components, and for the loadings themselves and residuals. Only the loadings and residuals are actually parameters (or something) but all of them have priors.

Update from the second meeting: Bob suggested using an ordination on the variance explained by the different latent variables, which I interpreted to be the genetic, not the total variance. The way he suggested implementing this was by using a prior on the additive genetic variance of the second prior being a constant times the first one, and this constant have a beta(1, 1) prior. This should, according to Bob, make it unnecessary to set the loading of a specific trait on one of the factors to zero, as the sampling would find the latent variable explaining the most variance automatically, and then find from this a second explaining less variance, and with this size relationship established, not have to wander aimlessly trying to figure out which factor is the "biggest". So this was attempted to be implemented, but for the first run I did set one loading to zero to be "safe" and see how that works, as well as setting one loading on each latent variable to 1, in order to fix the scale, which I am under the impression is necessary when having a free variance that is now no longer constrained to be 1, but which is still enforced to be one by having the z (LV) values standardized in a specific way Bob suggested, in which the mean of of the latent variable values are subtracted and then divided by the standard deviation, essentially. I also got sent an article that talks about an approach with an infinite number of factors that diminish based on the priors the higher its index it, which effectively works as parameter (loading) selection. This is the expanded version of the beta prior that Bob suggested. I might try this full approach later on.

Now this model has also failed to converge. So what I will try is to go back and implement one thing at a time, to make sure it works, and to see where the problem lies. First, I will check that the recursive approach with the Mendelian actually works, so I will go back to implementing a single latent variable model, only at the individual level, and with only 3 traits. Update: This worked, with tarsus as the first (anchor) trait, with a really high loading (close to 1) on that and a high heritability of around 0.4, but with low loadings on the two other traits (less than 0.2). However, when adding the two final traits (body mass and beak height), the model suddenly didn't converge. What I then did was very informative: I checked the correlations between the 5 traits in the data set, and saw that tarsus actually was not very correlated with several of the traits, but wing length was a much better anchor as it was correlated decently with all traits. This might mean that it is a better proxy for general body size. So I will run a model with that one as the first trait. In the 1 LV model the anchor trait works as the anchor by being defined to be positive. So essentially if the first trait doesn't correlate with the others, it just does its own thing, the model might struggle to distinguish between the factor not meaning anything at all and following the anchor trait, as it's not accounting for the other traits.

\% This is the table of the correlations: % % tars_h_std nebb_l_std ving_h_std vekt_std nebb_h_std % tars_h_std 1.0000000 0.1503248 0.2102723 0.2995245 0.1137911 % nebb_l_std 0.1503248 1.0000000 0.4018150 0.2892001 0.5862900 % ving_h_std 0.2102723 0.4018150 1.0000000 0.3196946 0.4148833 % vekt_std 0.2995245 0.2892001 0.3196946 1.0000000 0.3692893 % nebb_h_std 0.1137911 0.5862900 0.4148833 0.3692893 1.0000000

I wonder why I didn't check this earlier...

Update on the second run with wing length as the first trait: It did not work this time either. The conclusion to this is that the model does not accept the 1 latent variable reality for these 5 traits, as there seems to be one beak factor and one general size factor affecting the three other traits.

However, with the first anchor trait as the wing length and the second anchor as the tarsus, the model finally converged. This suggests that the tarsus is doing its own thing, separate from the other traits. However, this interpretation might be coloured by the fact that we are constraining the first loading on the second factor to zero. I will also try to swap the first two traits to see if I still get a similar interpretation. The results from this showed the same general result, although slightly different. In the first run, with wing length first, the second factor had some negative loadings, but in the second run, with tarsus first, all loadings were positive, as well as some other results being slightly different.

\% The correlation table (correlations of the matrix of actual data minus the means):

\% ving_h_std tars_h_std nebb_l_std nebb_h_std vekt_std % ving_h_std 1.0000000 0.2013426 0.2028627 0.4321624 0.3273366 % tars_h_std 0.2013426 1.0000000 0.1278506 0.1141831 0.3002953 % nebb_l_std 0.2028627 0.1278506 1.0000000 0.5055168 0.2573306 % nebb_h_std 0.4321624 0.1141831 0.5055168 1.0000000 0.3698353 % vekt_std 0.3273366 0.3002953 0.2573306 0.3698353 1.0000000

The correlation table suggests that there are lots of correlations not accounted for by the factors, suggesting either we need the temporary environmental layer or a third factor or something.

Next, something I should have done much earlier, I will remove outliers, as I realized there are likely some typos or other sources of error, causing some individuals to have very large quantile values (when standardizing the data). Something I also realized that I should have done is add the age as a fixed effect too, as there are many juveniles. This will also help to find outliers, as I could find the outliers from residuals after fitting a simple linear model with age and sex as fixed effects. This was done and the resulting plot shows the normal distribution of the residuals after accounting for sex and age.

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{Images/residuals_plot_sex_age.pdf}
    \caption{Resiudals plots (student t) of traits after sex and age effects are accounted for (fit lm)}
    \label{fig:resid_plot}
\end{figure}

The way I added the age as a fixed effect is the following way: There is a column called scriptage (with the values of the number of calendar years the birds have been alive) and a column called fieldage (with either juvenile or adult or unknown/probably determined in the field). I used the scriptsex column when available (which was for almost all birds), and if not I used fieldage (removing unknowns and mapping probably to the respective case). I only added 2 levels: For values of 1K or 1K+ I mapped it to level 1 (juvenile) and for values 2K, 2K+ or above I mapped it to level 2 (adult). I don't see a point in adding a level for each year, as we are not really not modelling ageing (yet), just whether or not the bird is full-grown.

I ran a model with the temporary environmental level with the constrained version of the model (first loading of second set to zero or what it is), and then I got the results. It first seemed it hadn't converged, but then I took a look at the histories of the chains, and it turned out that one chain was doing its own thing while the 3 others had converged to the same solution. Therefore, I removed the weird chain and got some interesting results. In this case the loading on both of the factors was stronger for the tarsus than the wing, which was different from when I didn't have the TE layer. By the way, it took a long time to run. This implies that maybe tarsus is in fact the best trait for the first loading, it has the strongest genetic basis in any case, and might be the closest related to a structural or bone-centred size concept which should be the first factor, so I will run it again with tarsus first, then wing second. However, I will try to implement the beta prior model first without any constraint, this time only with the individual level latent variable. But the standardisation and priors are a bit tricky here, and I wonder if it'll converge. But the improvement/changes could also have been due to the introduction of the age effect.

This time I ran the model with age effect included, which should also account for some variance previously unaccounted for, which could also explain the significantly reduced residual error for the tarsus (which was now almost zero). In any case the TE variance is not big at all compared to the other two components.

I also removed the most important outliers, but I realise now that I removed the entire bird (all measurements) if there was a single extreme measurement, when I should have just removed the single extreme measurement, I should fix this too, but all in all I didn't remove more than roughly 40 total birds out of more than 1,600.

I tried running a shrinkage prior model without constraining any loadings and without standardizing but with initializing, and I got interesting results. They suggest a size factor (loading strongest on tarsus) accounting for the largest variance, and then a shape factor with slightly less variance, which loaded positive on the two beak traits negative on the tarsus, suggesting a trade-off. The loadings were struggling to converge, but were not completely off, and the highest Rhat was 1.08, and I got neff under 100 for all the loadings, so I might run the model with at least twice the number of iterations or try something else, but it isn't a disaster run. The initialization was probably important, but I'm not entirely sure what made it find the right values without wobbling too much (although it did wobble). This suggests some ambiguity as to which factor should be leading, which is basically rotational ambiguity.

Essentially, there are three main issues in Bayesian factor analysis when it comes to identifiability: scale indeterminacy (large loading and low score or vice versa), rotational freedom and sign flipping. To remedy these issues, the script firstly defines the total variance of the first latent variable to be 1 by splitting it between the heritable part and the environmental, which each sum to 1. Then the rotational freedom issue is solved by the shrinkage prior on the loadings, forcing the second factor to be tighter, making sure that they are not interchangeable. The sign flipping is not explicitly solved for but is likely solved by the initialization.

Update: I ran this model, and it actually turned out that it produced sign flipping, but after correcting for that it produced nearly the exact same results, but with slightly higher neff and Rhats (lowest neff 90 and highest Rhat 1.04). This time I got the rho parameter estimated as well, which ended up between 0.09 and 0.7 (with mean 0.31) or so. Summing up the square of the loadings to try and estimate the amount of variance accounted for by each of the factors, I can take the square of the loadings, and that led to approximately 1.3 and 0.7. So that should correspond roughly to the square root of 0.31 times the loadings of the first latent variable (?) (look into this).

\% Here were the results after I did the sign flip:

\% mean sd 2.5% 97.5% n_eff Rhat % Lambda\[1,1\] 0.845 0.0833 0.6558 0.9710 115 1.03 % Lambda\[2,1\] 0.323 0.0430 0.2208 0.3867 102 1.03 % Lambda\[3,1\] 0.395 0.0851 0.1998 0.5303 93 1.03 % Lambda\[4,1\] 0.376 0.1006 0.1470 0.5402 92 1.04 % Lambda\[5,1\] 0.481 0.0588 0.3410 0.5665 98 1.03 % Lambda\[1,2\] -0.439 0.1745 -0.7296 -0.0403 90 1.04 % Lambda\[2,2\] 0.195 0.0658 0.0630 0.3235 99 1.03 % Lambda\[3,2\] 0.423 0.0797 0.2547 0.5673 102 1.04 % Lambda\[4,2\] 0.508 0.0758 0.3423 0.6358 109 1.03 % Lambda\[5,2\] 0.267 0.0967 0.0726 0.4576 96 1.04

Also, I performed a residual error analysis, and it seems the residuals are quite strongly correlated, based on a correlation table. Some correlations were as high in absolute value as 0.5. I did QQ plots and also residual plots (on predicted value). The latter seemed fine. The stripe patterns appear because some of the numbers are integers. The QQ plots showed that the tails were quite heavy, but were linear in the middle. This suggests (as we also saw from the outlier detection code) that there are more extreme outliers than one would expect if this were a true normal distribution. But I don't know if there's any point to trying a Student's t distribution, it's mostly normal. Maybe I'll ask Bob.

I also received the survival and reproduction data from Yimen, which needs to be taken into consideration.

so we will make 2 different "spokes". the model will estimate the lv's even still, just multiplying the likelihoods. they can both have a unique set of effects, both different fixed effects and random effects, and they are only linked together through the common latent factor (score). They can also have different link functions. this is multivariate GLMM with factor structure. on the morphological traits we want the sex, age, season (?), and year as random effect. for the reproduction we want year as random effect, as well as sex and age as fixed effects.

by the way, the best approach (which i'm doing) is to estimate also the fixed effects inside of Stan (in the sampling) to get credible intervals etc., as we don't know how much is fixed effect and how much is random effect/genetics etc. from the get go. we don't know the season in the reproduction data, but the beauty of this approach is we don't need it, it doesn't even make sense to have a season as these are annual data.

so steps: first just make a quick fixed effects + simple random effects model with both spokes, and look at the p values, lrt and residuals, before fitting the full combined model, i might even run a new morphological bayesian model with the new fixed effects before combining.

and yeah, the survival data is of no use to this model. and also, we can use the reproduction data as is and forget the total number of recruits issue as the number in the data is inherently a value of offspring per year.

```{r}
# ==============================================================================
# 1. SETUP & LIBRARIES
# ==============================================================================
personal_lib <- "~/R/library"
if (!dir.exists(personal_lib)) dir.create(personal_lib, recursive = TRUE)
.libPaths(c(personal_lib, .libPaths()))

required_packages <- c("rstan", "tidyverse", "nadiv", "MCMCglmm") 
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, lib = personal_lib, repos = "https://cloud.r-project.org")
    library(pkg, character.only = TRUE)
  }
}

# ==============================================================================
# 2. LOAD & CLEAN DATA (OPTIMIZED ORDER)
# ==============================================================================

# Pedigree
pedData <- read.csv("pedigree_data.txt", sep = " ", header = TRUE, na.strings = "NA")
pedigree_clean <- pedData %>%
  dplyr::select(ringnr, dam, sire) %>%
  mutate(across(c(ringnr, dam, sire), ~ as.character(trimws(gsub("_.*$", "", .)))),
         dam = if_else(is.na(dam) | dam == "" | dam == "NA", NA_character_, dam),
         sire = if_else(is.na(sire) | sire == "" | sire == "NA", NA_character_, sire))

# External Parents
all_parents <- unique(c(pedigree_clean$dam, pedigree_clean$sire))
external_parents <- setdiff(all_parents, c(pedigree_clean$ringnr, NA_character_))
if (length(external_parents) > 0) {
  founder_df <- data.frame(ringnr = external_parents, dam = NA_character_, sire = NA_character_)
  pedigree_full <- bind_rows(founder_df, pedigree_clean) %>% distinct(ringnr, .keep_all = TRUE)
} else {
  pedigree_full <- pedigree_clean
}

pedigree_prepped <- prepPed(pedigree_full, gender = NULL, check = TRUE)

# Trait Data
traitData <- read.csv("morphology.txt", sep = ";", header = TRUE, 
                      na.strings = c("NA", ""), fileEncoding = "Windows-1252", 
                      stringsAsFactors = FALSE, colClasses = c(ringnr = "character"))

# A. LOAD & LIGHT FORMATTING (Crucial for matching)
trait_raw <- traitData %>%
  filter(sted_r == "hestmannøy", 
         !is.na(ving_h), !is.na(nebb_l), !is.na(tars_h), !is.na(nebb_h), !is.na(vekt)) %>%
  mutate(ringnr = as.character(trimws(gsub("_.*$", "", ringnr))))

# B. ALIGN WITH PEDIGREE (The "Gatekeeper" Step)
valid_ids <- intersect(trait_raw$ringnr, pedigree_prepped$ringnr)
trait_subset_temp <- trait_raw %>% filter(ringnr %in% valid_ids)

cat("Birds in Raw File:    ", nrow(trait_raw), "\n")
cat("Birds with Pedigree:  ", nrow(trait_subset_temp), "\n")

# C. DEEP CLEANING (Sex, Age, Bad IDs) on the SUBSET
bad_ids <- c("8N27558", "8N42527", "8N13933", "8M71874", "8N87712", "8L89527")

trait_subset_temp <- trait_subset_temp %>%
  filter(!ringnr %in% bad_ids) %>%
  mutate(raw_sex = if_else(!is.na(scriptsex) & scriptsex != "u", scriptsex, fieldsex)) %>%
  mutate(sex = case_when(
    raw_sex %in% c("m", "pm") ~ "m", 
    raw_sex %in% c("f", "pf") ~ "f", 
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(sex)) %>%
  mutate(
    s_age = trimws(as.character(scriptage)),
    f_age = trimws(as.character(fieldage)),
    age_class = case_when(
      grepl("^1K", s_age) ~ "juvenile",
      grepl("^[2-9]K", s_age) | grepl("^1[0-9]K", s_age) ~ "adult",
      f_age %in% c("j", "pj") ~ "juvenile",
      f_age %in% c("a", "pa") ~ "adult",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_class))

# D. STATISTICAL OUTLIER REMOVAL (Residual Check)
outlier_ids_to_kill <- c()
traits_to_check <- c("ving_h", "tars_h", "nebb_l", "nebb_h", "vekt")

for(t in traits_to_check) {
  f <- as.formula(paste(t, "~ sex + age_class"))
  fit <- lm(f, data = trait_subset_temp, na.action = na.exclude)
  
  resids <- rstudent(fit)
  cutoff <- if(t == "vekt") 6.0 else 4.0 
  
  bad_rows <- which(abs(resids) > cutoff)
  
  if(length(bad_rows) > 0) {
    new_bad_ids <- trait_subset_temp$ringnr[bad_rows]
    outlier_ids_to_kill <- c(outlier_ids_to_kill, new_bad_ids)
    cat("Trait:", t, "- Removing", length(new_bad_ids), "IDs\n")
  }
}

# E. FINAL DATASET
trait_subset_lv <- trait_subset_temp %>%
  filter(!ringnr %in% unique(outlier_ids_to_kill))

cat("Final Clean N: ", nrow(trait_subset_lv), "\n")



# NEW: Log-transform FIRST, then scale
trait_subset_lv <- trait_subset_lv %>%
  mutate(
    ving_h_std = as.numeric(scale(log(ving_h))),
    nebb_l_std = as.numeric(scale(log(nebb_l))),
    tars_h_std = as.numeric(scale(log(tars_h))),
    nebb_h_std = as.numeric(scale(log(nebb_h))),
    vekt_std   = as.numeric(scale(log(vekt)))
  )


cat("Data Loaded. N Rows:", nrow(trait_subset_lv), "\n")




# 1. Convert to Date (R handles YYYY-MM-DD automatically)
trait_subset_lv$date_obj <- as.Date(trait_subset_lv$dato)

# 2. Extract month
trait_subset_lv$month <- as.numeric(format(trait_subset_lv$date_obj, "%m"))



# 1. Extract the Year as a factor
trait_subset_lv <- trait_subset_lv %>%
  mutate(
    Year = factor(format(date_obj, "%Y")),
    # Create the 3-level Season we discussed
    Season_3 = factor(case_when(
      month %in% c(5, 6, 7) ~ "Breeding",
      month %in% c(8, 9)    ~ "Molt",
      TRUE                  ~ "Winter"
    ))
  )






# Grouping low-volume measurers using case_when
trait_subset_lv <- trait_subset_lv %>%
  group_by(init) %>%
  mutate(init_clean = case_when(
    n() < 10 ~ "OTHER",
    TRUE     ~ as.character(init)
  )) %>%
  ungroup() %>%
  mutate(init_clean = factor(init_clean))


###############################################################

############### DID I NOT ADD YEAR HERE?????? ##################

##############################################################

############# =============================================


# ==============================================================================
# 2.5 LOAD & CLEAN FITNESS DATA
# ==============================================================================

fitness_data <- read.csv("Fitness.csv", header = TRUE, stringsAsFactors = FALSE, na.strings = "NA", fileEncoding = "Windows-1252")

fitness_data_clean <- fitness_data %>%
  dplyr::select(Location, Year_ID, Year, Sex, Own_survival, N_Recruits, Least_age) %>%
  filter(grepl("^Hestmann", Location)) %>%
  mutate(
    # Extract the ring number
    ringnr = sub("^.*_", "", Year_ID),
    
    # NEW: Extract the actual calendar year and overwrite the 1-22 index
    Year = sub("_.*$", "", Year_ID), 
    
    Own_survival = as.numeric(Own_survival)
  ) %>%
  # Gatekeeper: Must be in the prepared pedigree
  filter(ringnr %in% pedigree_prepped$ringnr)

# Split into two clean datasets (Stan cannot handle NA in the response variable)
data_repro <- fitness_data_clean %>% drop_na(N_Recruits)
data_surv  <- fitness_data_clean %>% drop_na(Own_survival)

cat("Morphology records: ", nrow(trait_subset_lv), "\n")
cat("Reproduction records: ", nrow(data_repro), "\n")
cat("Survival records: ", nrow(data_surv), "\n")







# ==============================================================================
# 3. PEDIGREE PROCESSING
# ==============================================================================
pedigree_working <- pedigree_prepped
colnames(pedigree_working) <- c("id", "dam", "sire") 
pedigree_working$id   <- as.character(pedigree_working$id)
pedigree_working$dam  <- as.character(pedigree_working$dam)
pedigree_working$sire <- as.character(pedigree_working$sire)

# NEW: Keep all birds that appear in ANY of the three datasets
birds_to_keep <- unique(c(trait_subset_lv$ringnr, data_repro$ringnr, data_surv$ringnr))

ped_pruned <- MCMCglmm::prunePed(pedigree_working, keep = birds_to_keep)

A_inv_obj <- MCMCglmm::inverseA(ped_pruned, nodes = "ALL", scale = FALSE)
ordered_ids <- rownames(A_inv_obj$Ainv)

ped_ordered <- data.frame(
  id = ordered_ids,
  dam = as.character(A_inv_obj$pedigree[, 2]), 
  sire = as.character(A_inv_obj$pedigree[, 3]),
  stringsAsFactors = FALSE
)

dii_vector <- A_inv_obj$dii
Na <- nrow(ped_ordered)

dam_idx_mapped  <- match(ped_ordered$dam, ped_ordered$id)
sire_idx_mapped <- match(ped_ordered$sire, ped_ordered$id)

dam_idx_mapped[is.na(dam_idx_mapped)]   <- Na + 1
sire_idx_mapped[is.na(sire_idx_mapped)] <- Na + 1

id_map <- setNames(1:Na, ped_ordered$id)






# ==============================================================================
# 4. DATA MAPPING & UNIVERSAL FACTORS
# ==============================================================================
# 1. Map IDs
animal_morph <- unname(id_map[trait_subset_lv$ringnr])
animal_repro <- unname(id_map[data_repro$ringnr])
animal_surv  <- unname(id_map[data_surv$ringnr])

# 2. Universal Year Map (Crucial so Year 1 is the same everywhere)
all_years <- sort(unique(c(as.character(trait_subset_lv$Year), 
                           as.character(data_repro$Year), 
                           as.character(data_surv$Year))))
year_map <- setNames(1:length(all_years), all_years)

year_morph <- unname(year_map[as.character(trait_subset_lv$Year)])
year_repro <- unname(year_map[as.character(data_repro$Year)])
year_surv  <- unname(year_map[as.character(data_surv$Year)])

# 3. Measurer Map
init_factor <- as.integer(factor(trait_subset_lv$init_clean))

# 4. Trait Matrix
Y_mat <- as.matrix(trait_subset_lv[, c("tars_h_std", "ving_h_std", "nebb_l_std", "nebb_h_std", "vekt_std")])




# ==============================================================================
# 5. STAN EXECUTION FOR JOINT MODEL
# ==============================================================================
dataset_joint <- list(
  # Dimensions
  No_morph = nrow(trait_subset_lv),
  No_repro = nrow(data_repro),
  No_surv  = nrow(data_surv),
  Nt = 5, 
  Nlv = 1,
  Na = Na,
  
  # Responses
  Y_morph = Y_mat,
  Y_repro = as.array(as.integer(data_repro$N_Recruits)),
  Y_surv  = as.array(as.integer(data_surv$Own_survival)),
  
  # Animal Indices
  animal_morph = as.array(as.integer(animal_morph)),
  animal_repro = as.array(as.integer(animal_repro)),
  animal_surv  = as.array(as.integer(animal_surv)),
  
  # Pedigree
  dam  = as.array(as.integer(dam_idx_mapped)), 
  sire = as.array(as.integer(sire_idx_mapped)),
  dii  = as.array(dii_vector),
  
  # Random Effects Indices
  N_year = length(all_years),
  year_morph = as.array(as.integer(year_morph)),
  year_repro = as.array(as.integer(year_repro)),
  year_surv  = as.array(as.integer(year_surv)),
  
  N_init = max(init_factor),
  init_morph = as.array(as.integer(init_factor))
)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())


init_fn_joint <- function() {
  list(
    # Core Morphology (Anchored for 1 LV)
    lambda_anchor = array(runif(dataset_joint$Nlv, 0.5, 1.2)), 
    lambda_free = matrix(rnorm((dataset_joint$Nt - 1) * dataset_joint$Nlv, 0, 0.1), 
                         nrow = dataset_joint$Nt - 1, ncol = dataset_joint$Nlv),
    rho = runif(1, 0.4, 0.6),        
    h2_lv = array(runif(dataset_joint$Nlv, 0.4, 0.6)),      
    sd_R = runif(dataset_joint$Nt, 0.5, 1.0), 
    
    # Intercepts & Slopes (Fitness gradients as arrays of length 1)
    alpha_trait = rnorm(dataset_joint$Nt, 0, 0.5),
    alpha_repro = rnorm(1, -0.5, 0.5),
    alpha_surv  = rnorm(1, 0, 0.5),
    gamma_repro = array(rnorm(dataset_joint$Nlv, 0, 0.1)),
    gamma_surv  = array(rnorm(dataset_joint$Nlv, 0, 0.1)),
    
    # Genetic Noise (Now a 1 x Na matrix)
    w_a = matrix(rnorm(dataset_joint$Nlv * dataset_joint$Na, 0, 0.1), 
                 nrow = dataset_joint$Nlv),
    w_pe = matrix(rnorm(dataset_joint$Nlv * dataset_joint$Na, 0, 0.1), 
                  nrow = dataset_joint$Nlv),
    
    # Random Effects
    sd_year_morph = runif(dataset_joint$Nt, 0.05, 0.2),
    sd_year_repro = runif(1, 0.05, 0.2),
    sd_year_surv  = runif(1, 0.05, 0.2),
    sd_init_morph = runif(dataset_joint$Nt, 0.05, 0.2),
    
    z_year_morph = matrix(rnorm(dataset_joint$N_year * dataset_joint$Nt, 0, 0.1), 
                          nrow = dataset_joint$N_year),
    z_year_repro = rnorm(dataset_joint$N_year, 0, 0.1),
    z_year_surv  = rnorm(dataset_joint$N_year, 0, 0.1),
    z_init_morph = matrix(rnorm(dataset_joint$N_init * dataset_joint$Nt, 0, 0.1), 
                          nrow = dataset_joint$N_init)
  )
}

# When calling stan(), remember to use this updated init:
# init = init_fn_joint_1lv




# Run the Joint Model
out_joint <- stan(
  file = "combined_model_test.stan",
  data = dataset_joint,
  init = init_fn_joint,        
  chains = 4, 
  pars = c("alpha_trait", "alpha_repro", "alpha_surv", 
           "gamma_repro", "gamma_surv", "Lambda", "sd_R", 
           "sd_year_morph", "sd_year_repro", "sd_year_surv", "sd_init_morph", 
           "h2_lv"
           # , "rho"
           ),
  control = list(adapt_delta = 0.95), 
  iter = 6000,          
  warmup = 2000,
  sample_file = "output_joint_chain"
)

saveRDS(out_joint, 'Output_Joint_Starter.rds')

















# ==============================================================================
# 6. RESULTS & DIAGNOSTICS
# ==============================================================================
out_lv <- readRDS('output_markov/Output_Joint_Starter.rds')

# Print the parameters specific to the Joint Starter model
print(out_lv,
      pars = c("alpha_trait",       # Base intercepts for the 5 morph traits
               "alpha_repro",       # Base intercept for reproduction
               "alpha_surv",        # Base intercept for survival
               "gamma_repro",       # Selection gradient: Size -> Reproduction!
               "gamma_surv",        # Selection gradient: Size -> Survival!
               "Lambda",            # Factor loadings (How traits build Size)
               "h2_lv",             # Heritability of the Latent Size
               # "rho",               # Shrinkage factor
               "sd_R",              # Residual error for morph traits
               "sd_year_morph",     # Year variance for morphology
               "sd_year_repro",     # Year variance for reproduction
               "sd_year_surv",      # Year variance for survival
               "sd_init_morph",     # Measurer variance for morphology
               "lp__"),             # Log-posterior (model fit)
      probs = c(0.025, 0.975),
      digits_summary = 3)





library(bayesplot)
library(ggplot2)

# 1. Plot a stable parameter (The Grand Mean)
color_scheme_set("mix-blue-red")
p1 <- mcmc_trace(out_lv, pars = c("alpha_trait[1]", "sd_year_surv")) +
  ggtitle("Stable Parameters (Rhat = 1.00)")

# 2. Plot the 'Flipped' parameters
p2 <- mcmc_trace(out_lv, pars = c("Lambda[1,1]", "gamma_repro[1]", "gamma_surv[1]")) +
  ggtitle("Sign-Flipped Parameters (High Rhat)")

p3 <- mcmc_trace(out_lv, pars = c("Lambda[1,2]", "Lambda[3,2]", "gamma_repro[2]", "gamma_surv[2]"))
  

# Show the plots
print(p1)
print(p2)
print(p3)







```

A quick check here showed that there were quite a few birds that were phenotyped but that were not in the fitness data, so that we are missing about 30 percent of the measurements if we only select birds that are in all data sets. So, the solution here might be to keep all phenotyped birds, but only have fitness data for the ones that have that, and that is possible to do as far as I know. (Turns out that's what we're doing, all birds in the pedigree are taken into account, but only the information available at each row of data (be it fitness data or morphology data) is used, and we don't need all data at each row or for each bird for it to work).

```{stan}
data {
  // Dimensions
  int<lower=1> No_morph; 
  int<lower=1> No_repro;
  int<lower=1> No_surv;
  int<lower=1> Nt;  
  int<lower=1> Nlv; 
  int<lower=1> Na; 
  
  // Responses
  matrix[No_morph, Nt] Y_morph; 
  array[No_repro] int<lower=0> Y_repro; // Poisson count
  array[No_surv] int<lower=0, upper=1> Y_surv; // Bernoulli 0/1
  
  // Pedigree & Links
  array[No_morph] int<lower=1, upper=Na> animal_morph; 
  array[No_repro] int<lower=1, upper=Na> animal_repro;
  array[No_surv]  int<lower=1, upper=Na> animal_surv;
  
  array[Na] int<lower=1, upper=Na+1> dam; 
  array[Na] int<lower=1, upper=Na+1> sire;
  vector[Na] dii; 
  
  // Random Effects Indices
  int<lower=1> N_year;
  array[No_morph] int<lower=1, upper=N_year> year_morph;
  array[No_repro] int<lower=1, upper=N_year> year_repro;
  array[No_surv]  int<lower=1, upper=N_year> year_surv;
  
  int<lower=1> N_init;
  array[No_morph] int<lower=1, upper=N_init> init_morph;
}

parameters {
  // Intercepts (Since we dropped the X matrix)
  vector[Nt] alpha_trait;
  real alpha_repro;
  real alpha_surv;

  // Selection Gradients (The link between Size and Fitness!)
  vector[Nlv] gamma_repro;
  vector[Nlv] gamma_surv;

  // Variances
  vector<lower=0>[Nt] sd_R; 
  // real<lower=0, upper=1> rho; 
  vector<lower=0, upper=1>[Nlv] h2_lv; 
  matrix[Nt, Nlv] lambda_raw;

  // Latent Effects
  matrix[Nlv, Na] w_a;  
  matrix[Nlv, Na] w_pe; 
  
  // Random Effects (Year and Measurer)
  matrix[N_year, Nt] z_year_morph;
  vector<lower=0>[Nt] sd_year_morph;
  
  vector[N_year] z_year_repro;
  real<lower=0> sd_year_repro;
  
  vector[N_year] z_year_surv;
  real<lower=0> sd_year_surv;
  
  matrix[N_init, Nt] z_init_morph;
  vector<lower=0>[Nt] sd_init_morph;
}

transformed parameters {
  // 1. Pedigree Recursion
  matrix[Na, Nlv] G_effect; 
  {
    matrix[Na+1, Nlv] aug_a;
    aug_a[Na+1] = rep_row_vector(0, Nlv);
    for (i in 1:Na) {
       aug_a[i] = 0.5 * aug_a[dam[i]] + 0.5 * aug_a[sire[i]] 
                  + sqrt(dii[i]) * w_a[, i]'; 
    }
    G_effect = aug_a[1:Na];
  }

  // 2. Latent Factor
  matrix[Na, Nlv] LV;
  for(k in 1:Nlv) {
    LV[, k] = sqrt(h2_lv[k]) * G_effect[, k] + sqrt(1 - h2_lv[k]) * w_pe[k, ]';
  }

  // 3. Shrinkage
  matrix[Nt, Nlv] Lambda;
  Lambda[, 1] = lambda_raw[, 1];            
  // Lambda[, 2] = lambda_raw[, 2] * sqrt(rho); 
}

model {
  // Priors
  // rho ~ beta(1.7, 2.5);
  h2_lv ~ beta(2, 2); 
  to_vector(lambda_raw) ~ std_normal();
  sd_R ~ normal(0, 1);
  
  alpha_trait ~ normal(0, 2);
  alpha_repro ~ normal(0, 2);
  alpha_surv ~ normal(0, 2);
  
  gamma_repro ~ normal(0, 1);
  gamma_surv ~ normal(0, 1);

  to_vector(w_a) ~ std_normal();
  to_vector(w_pe) ~ std_normal();
  to_vector(z_year_morph) ~ std_normal();
  z_year_repro ~ std_normal();
  z_year_surv ~ std_normal();
  to_vector(z_init_morph) ~ std_normal();
  
  sd_year_morph ~ exponential(2); 
  sd_year_repro ~ exponential(2); 
  sd_year_surv ~ exponential(2); 
  sd_init_morph ~ exponential(2);

  // --- LIKELIHOODS ---
  
  // 1. Morphology
  for(i in 1:No_morph){
    for(t in 1:Nt){
      real mu = alpha_trait[t] 
                + LV[animal_morph[i], 1] * Lambda[t, 1] 
                // + LV[animal_morph[i], 2] * Lambda[t, 2]
                + z_year_morph[year_morph[i], t] * sd_year_morph[t]
                + z_init_morph[init_morph[i], t] * sd_init_morph[t];
      Y_morph[i, t] ~ normal(mu, sd_R[t]);
    }
  }
  

  for(i in 1:No_repro){
    real log_lambda = alpha_repro 
                      + gamma_repro[1] * LV[animal_repro[i], 1] 
                      + z_year_repro[year_repro[i]] * sd_year_repro;
    Y_repro[i] ~ poisson_log(log_lambda);
  }
  
  // 3. Survival (Bernoulli)
  for(i in 1:No_surv){
    real logit_p = alpha_surv 
                   + gamma_surv[1] * LV[animal_surv[i], 1] 
                   + z_year_surv[year_surv[i]] * sd_year_surv;
    Y_surv[i] ~ bernoulli_logit(logit_p);
  }
}



```

This run with 1 latent variable on the 5 traits, with survival and reproduction added led to these successful results (this was without all the fixed and random effects by the way):

::: output
Chain 4: 16775.2 seconds (Total) Chain 4: Inference for Stan model: anon_model. 4 chains, each with iter=6000; warmup=2000; thin=1; post-warmup draws per chain=4000, total post-warmup draws=16000.

```         
                  mean se_mean     sd      2.5%     97.5% n_eff  Rhat
```

alpha_trait\[1\] -0.022 0.001 0.057 -0.131 0.091 4522 1.001 alpha_trait\[2\] 0.037 0.001 0.064 -0.089 0.162 4292 1.001 alpha_trait\[3\] 0.022 0.001 0.065 -0.103 0.150 3924 1.001 alpha_trait\[4\] 0.202 0.001 0.079 0.050 0.361 3457 1.000 alpha_trait\[5\] 0.111 0.001 0.071 -0.027 0.250 3974 1.001 alpha_repro -0.354 0.002 0.100 -0.553 -0.156 2557 1.001 alpha_surv 0.186 0.001 0.085 0.015 0.354 6752 1.000 gamma_repro\[1\] 0.074 0.000 0.024 0.026 0.122 30770 1.000 gamma_surv\[1\] 0.115 0.000 0.042 0.033 0.196 30325 1.000 Lambda\[1,1\] 0.970 0.000 0.020 0.931 1.011 2055 1.003 Lambda\[2,1\] 0.226 0.000 0.015 0.196 0.255 13184 1.000 Lambda\[3,1\] 0.158 0.000 0.015 0.129 0.187 21086 1.000 Lambda\[4,1\] 0.119 0.000 0.015 0.091 0.149 24667 1.000 Lambda\[5,1\] 0.316 0.000 0.016 0.286 0.347 8836 1.001 h2_lv\[1\] 0.472 0.001 0.046 0.382 0.559 1366 1.003 sd_R\[1\] 0.189 0.000 0.003 0.184 0.194 9801 1.000 sd_R\[2\] 0.905 0.000 0.010 0.886 0.924 35016 1.000 sd_R\[3\] 0.916 0.000 0.010 0.896 0.935 31259 1.000 sd_R\[4\] 0.902 0.000 0.010 0.883 0.921 33600 1.000 sd_R\[5\] 0.891 0.000 0.010 0.872 0.910 37002 1.000 sd_year_morph\[1\] 0.019 0.000 0.010 0.001 0.042 3976 1.002 sd_year_morph\[2\] 0.207 0.001 0.041 0.139 0.297 5018 1.001 sd_year_morph\[3\] 0.173 0.000 0.034 0.116 0.250 6459 1.000 sd_year_morph\[4\] 0.262 0.001 0.054 0.171 0.382 4735 1.000 sd_year_morph\[5\] 0.254 0.001 0.049 0.172 0.365 4871 1.001 sd_year_repro 0.443 0.001 0.078 0.316 0.621 4296 1.001 sd_year_surv 0.344 0.001 0.069 0.232 0.505 6477 1.000 sd_init_morph\[1\] 0.160 0.000 0.019 0.128 0.203 3337 1.000 sd_init_morph\[2\] 0.278 0.001 0.039 0.210 0.365 4960 1.001 sd_init_morph\[3\] 0.322 0.001 0.046 0.243 0.422 4875 1.000 sd_init_morph\[4\] 0.341 0.001 0.046 0.261 0.442 4573 1.000 sd_init_morph\[5\] 0.275 0.001 0.042 0.204 0.365 4956 1.001 lp\_\_ -8920.553 1.288 68.006 -9055.994 -8789.629 2786 1.002

Samples were drawn using NUTS(diag_e) at Fri Feb 27 15:52:35 2026. For each parameter, n_eff is a crude measure of effective sample size, and Rhat is the potential scale reduction factor on split chains (at convergence, Rhat=1).
:::

By the way, it turns out we are removing approximately 61 rows (out of a total of 4424) that do not have a sex value. So the question is whether it is worth it to include them (by finding the sex from a different place)?

```{r}
# ==============================================================================
# 6. RESULTS & DIAGNOSTICS
# ==============================================================================
# out_lv <- readRDS('Output_everything_model.rds')

# Print the parameters specific to the Joint Starter model
print(out_lv,
      pars = c("beta_morph", "beta_repro", "beta_surv",
               "gamma_repro", "gamma_surv", "Lambda", "sd_R", 
               "sd_year_morph", "sd_year_repro", "sd_year_surv", "sd_init_morph", 
               "h2_lv",
               "rho",    # Measurer variance for morphology
               "lp__"),             # Log-posterior (model fit)
      probs = c(0.025, 0.975),
      digits_summary = 3)




# Plot the main loading matrix and shrinkage
library(rstan)

# Use your actual fit object name (e.g., out_joint)
stan_trace(out_lv, pars = c("Lambda", "rho"))

# Look specifically at the Selection Gradients for the 2nd LV
stan_trace(out_lv, pars = c("gamma_repro", "gamma_surv"))

# Plot the main loading matrix and shrinkage

stan_trace(out_lv, pars = c("beta_morph[2,3]", "beta_morph[2,4]", "beta_morph[2,5]", 
                            "beta_morph[3,3]", "beta_morph[3,4]", "beta_morph[3,5]",
                            "beta_morph[4,3]", "beta_morph[4,4]", "beta_morph[4,5]"))



```

Alright, so it might be an issue that both age and season (as fixed effects) are struggling against the latent variables for being the source of variance. In any case it seems for the most part that also in the case of the beta's not converging it's mainly chain 4's fault.

Below are the results of the full run with a bunch of both fixed and random effects and also reproduction and survival added. It failed. Notice that also some of the beta's are not converged. This could be the result of a "tug-of-war" between the fixed effects and the latent variables, for example age is a fixed effect, but that also accounts for variation in general body size, so having that independent of the LV causes the model to not understand where to put the variation, and that could lead to convergence issues.

Inference for Stan model: anon_model. 4 chains, each with iter=6000; warmup=2000; thin=1; post-warmup draws per chain=4000, total post-warmup draws=16000.

```         
                  mean se_mean     sd      2.5%     97.5% n_eff  Rhat
```

beta_morph\[1,1\] -0.032 0.001 0.057 -0.142 0.081 7807 1.000 beta_morph\[1,2\] 0.208 0.001 0.065 0.081 0.338 4268 1.006 beta_morph\[1,3\] 0.292 0.001 0.058 0.177 0.407 5281 1.003 beta_morph\[1,4\] 0.347 0.003 0.077 0.197 0.498 908 1.014 beta_morph\[1,5\] 0.065 0.001 0.068 -0.067 0.199 5037 1.005 beta_morph\[2,1\] -0.002 0.000 0.010 -0.022 0.017 21881 1.001 beta_morph\[2,2\] -0.828 0.000 0.026 -0.879 -0.776 6707 1.005 beta_morph\[2,3\] -0.946 0.021 0.040 -1.016 -0.863 4 1.423 beta_morph\[2,4\] -0.927 0.023 0.042 -0.997 -0.839 3 1.513 beta_morph\[2,5\] -0.304 0.013 0.036 -0.373 -0.233 8 1.142 beta_morph\[3,1\] 0.040 0.000 0.012 0.017 0.063 23049 1.000 beta_morph\[3,2\] -0.011 0.000 0.033 -0.075 0.054 10614 1.004 beta_morph\[3,3\] -0.013 0.006 0.039 -0.090 0.065 46 1.028 beta_morph\[3,4\] 0.261 0.006 0.038 0.187 0.337 47 1.027 beta_morph\[3,5\] 0.250 0.001 0.043 0.166 0.333 1458 1.007 beta_morph\[4,1\] -0.005 0.000 0.012 -0.028 0.018 21404 1.000 beta_morph\[4,2\] 0.402 0.001 0.033 0.336 0.466 571 1.011 beta_morph\[4,3\] 0.107 0.015 0.044 0.021 0.193 8 1.139 beta_morph\[4,4\] 0.344 0.018 0.045 0.257 0.429 7 1.186 beta_morph\[4,5\] 0.321 0.009 0.045 0.232 0.409 23 1.050 beta_repro\[1\] -0.186 0.002 0.100 -0.388 0.008 3870 1.000 beta_repro\[2\] -0.411 0.000 0.048 -0.506 -0.315 27105 1.000 beta_surv\[1\] 0.096 0.001 0.094 -0.087 0.283 9311 1.000 beta_surv\[2\] 0.193 0.000 0.081 0.035 0.352 26845 1.000 gamma_repro\[1\] 0.059 0.000 0.025 0.011 0.107 22693 1.000 gamma_repro\[2\] 0.042 0.008 0.033 -0.024 0.104 18 1.063 gamma_surv\[1\] 0.123 0.000 0.043 0.038 0.206 10126 1.000 gamma_surv\[2\] 0.122 0.012 0.053 0.022 0.229 20 1.057 Lambda\[1,1\] 0.969 0.000 0.020 0.930 1.010 3029 1.002 Lambda\[1,2\] -0.018 0.002 0.064 -0.151 0.109 874 1.002 Lambda\[2,1\] 0.226 0.001 0.041 0.146 0.307 1126 1.005 Lambda\[2,2\] 0.589 0.071 0.103 0.378 0.689 2 4.458 Lambda\[3,1\] 0.166 0.001 0.028 0.109 0.225 747 1.007 Lambda\[3,2\] 0.307 0.070 0.100 0.210 0.508 2 4.697 Lambda\[4,1\] 0.121 0.001 0.031 0.058 0.187 743 1.007 Lambda\[4,2\] 0.361 0.074 0.106 0.262 0.569 2 5.080 Lambda\[5,1\] 0.311 0.001 0.026 0.258 0.363 983 1.005 Lambda\[5,2\] 0.297 0.034 0.055 0.221 0.420 3 2.065 sd_R\[1\] 0.187 0.000 0.003 0.182 0.192 9389 1.000 sd_R\[2\] 0.586 0.058 0.083 0.518 0.741 2 7.307 sd_R\[3\] 0.752 0.026 0.037 0.677 0.789 2 4.158 sd_R\[4\] 0.726 0.032 0.046 0.635 0.771 2 5.081 sd_R\[5\] 0.835 0.009 0.015 0.802 0.859 3 1.651 sd_year_morph\[1\] 0.016 0.000 0.010 0.001 0.038 4992 1.000 sd_year_morph\[2\] 0.128 0.001 0.035 0.072 0.207 3849 1.006 sd_year_morph\[3\] 0.125 0.003 0.028 0.077 0.188 91 1.022 sd_year_morph\[4\] 0.234 0.013 0.059 0.139 0.369 20 1.067 sd_year_morph\[5\] 0.193 0.002 0.042 0.123 0.287 388 1.015 sd_year_repro 0.442 0.001 0.076 0.315 0.610 5345 1.001 sd_year_surv 0.347 0.001 0.070 0.232 0.503 7126 1.000 sd_init_morph\[1\] 0.156 0.000 0.019 0.124 0.198 4464 1.000 sd_init_morph\[2\] 0.265 0.001 0.035 0.206 0.343 2671 1.008 sd_init_morph\[3\] 0.278 0.001 0.039 0.211 0.363 1323 1.011 sd_init_morph\[4\] 0.318 0.001 0.041 0.247 0.409 5754 1.001 sd_init_morph\[5\] 0.268 0.001 0.043 0.193 0.361 5383 1.005 h2_lv\[1\] 0.471 0.001 0.046 0.381 0.560 1989 1.001 h2_lv\[2\] 0.542 0.016 0.065 0.415 0.670 16 1.093 rho 0.158 0.030 0.113 0.031 0.456 14 1.084 lp\_\_ -7583.915 1.492 91.523 -7765.291 -7409.708 3762 1.003

Samples were drawn using NUTS(diag_e) at Thu Mar 5 05:51:39 2026. For each parameter, n_eff is a crude measure of effective sample size, and Rhat is the potential scale reduction factor on split chains (at convergence, Rhat=1).

Now testing what happens if I remove the one chain that disagreed a lot with the others, chain 4. But that isn't really good practice to remove one chain after the fact, but yeah:

```{r}
library(rstan)



post_array <- as.array(out_lv)

 # 2. Subset the array to keep only chains 2, 3, and 4
 # The second dimension of the array is the chain index
 array_subset <- post_array[, 1:3, ]
 
 
 new_summary <- monitor(array_subset, digits_summary = 3, print = FALSE)
 
 # 2. Convert to a data frame or matrix to enable row-subsetting
 new_summary_df <- as.data.frame(new_summary)
 
 

 # 3. Use rstan's monitor function to generate a new summary table
 # This will recalculate Mean, SD, n_eff, and R-hat for the remaining chains
 
 # 4. View the specific parameters that were "messy" before
 pars_to_check <- c("gamma_repro[1]", "gamma_repro[2]", 
                    "gamma_surv[1]", "gamma_surv[2]", 
                    "Lambda[1,1]", "Lambda[1,2]", 
                    "Lambda[2,1]", "Lambda[2,2]", 
                    "rho", "h2_lv[2]")
 
 # 3. Print your parameters of interest
 print(new_summary_df[pars_to_check, ])
```

```         
                  mean      se_mean         sd         2.5%         25%         50%        75%
```

gamma_repro\[1\] 0.05876032 0.0002661359 0.02524820 0.009481862 0.04175122 0.05915998 0.07640587 gamma_repro\[2\] 0.04777510 0.0003172129 0.02986328 -0.010871722 0.02730378 0.04818482 0.06733020 gamma_surv\[1\] 0.12208226 0.0004901230 0.04279028 0.037763686 0.09338545 0.12184798 0.15100881 gamma_surv\[2\] 0.11208502 0.0005623019 0.04955137 0.014866939 0.07866373 0.11199741 0.14616805 Lambda\[1,1\] 0.96976553 0.0005421117 0.02003019 0.931859582 0.95627350 0.96941554 0.98287756 Lambda\[1,2\] -0.01921731 0.0023323181 0.05477292 -0.142738853 -0.04933346 -0.01567230 0.01635498 Lambda\[2,1\] 0.22380085 0.0016100168 0.03987824 0.149546045 0.19797763 0.22217643 0.24814697 Lambda\[2,2\] 0.64693420 0.0007073727 0.02299174 0.600572777 0.63206784 0.64742157 0.66219629 rho 0.13362541 0.0023481142 0.09816247 0.027553851 0.06450410 0.10327557 0.17425838 h2_lv\[2\] 0.55484251 0.0023513491 0.06180481 0.434449028 0.51157360 0.55515390 0.59773809 97.5% n_eff Rhat valid Q5 Q50 Q95 MCSE_Q2.5 gamma_repro\[1\] 0.10700486 8802 1.0004085 1 0.0163985275 0.05915998 0.09930538 0.0009446112 gamma_repro\[2\] 0.10743334 8826 1.0002299 1 -0.0009635907 0.04818482 0.09638510 0.0010513245 gamma_surv\[1\] 0.20544156 7597 1.0002403 1 0.0525235148 0.12184798 0.19318054 0.0017109337 gamma_surv\[2\] 0.20727535 7694 0.9998258 1 0.0305582623 0.11199741 0.19381430 0.0023681569 Lambda\[1,1\] 1.01051147 1352 1.0011862 1 0.9374574698 0.96941554 1.00355302 0.0009286183 Lambda\[1,2\] 0.07991625 550 1.0040894 1 -0.1159916319 -0.01567230 0.06274527 0.0065584238 Lambda\[2,1\] 0.30945131 612 1.0029961 1 0.1610302094 0.22217643 0.29281746 0.0025564566 Lambda\[2,2\] 0.69168214 1046 1.0016785 1 0.6091233641 0.64742157 0.68412126 0.0020906611 rho 0.40826256 1734 1.0013450 1 0.0337616989 0.10327557 0.34081387 0.0010535992 h2_lv\[2\] 0.67290882 682 1.0017496 1 0.4520669502 0.55515390 0.65596094 0.0028583820 MCSE_Q25 MCSE_Q50 MCSE_Q75 MCSE_Q97.5 MCSE_SD Bulk_ESS Tail_ESS gamma_repro\[1\] 0.0004963538 0.0004024525 0.0004945841 0.0009398345 0.0002032631 8982 4495 gamma_repro\[2\] 0.0005834702 0.0003947184 0.0005480688 0.0014949426 0.0002697446 8902 4460 gamma_surv\[1\] 0.0009158824 0.0005036031 0.0007810828 0.0013057482 0.0003705827 7607 4188 gamma_surv\[2\] 0.0008147991 0.0007712307 0.0010868857 0.0016850624 0.0004293244 7771 4687 Lambda\[1,1\] 0.0005748420 0.0005895095 0.0005358934 0.0011162593 0.0003834128 1365 3118 Lambda\[1,2\] 0.0026418117 0.0018097028 0.0018903110 0.0045931769 0.0016500709 575 739 Lambda\[2,1\] 0.0014175238 0.0013694119 0.0018220951 0.0048068249 0.0011508057 627 915 Lambda\[2,2\] 0.0008354725 0.0007201968 0.0007976987 0.0011090193 0.0005003262 1065 1768 rho 0.0013674191 0.0019084002 0.0035337562 0.0085709880 0.0016606446 1616 2778 h2_lv\[2\] 0.0029986225 0.0024574171 0.0029831721 0.0033503018 0.0016671106 692 1601

So much better results here. But should probably check the log posterior of the different chains. And can't just remove a chain I don't like.

```{r}
# 1. Extract the 3D array from your Stan fit
post_array <- as.array(out_lv)

# 2. Define a quick function to get the Mean and 95% CI
summarize_chain <- function(x) {
  c(Mean = mean(x), 
    `2.5%` = quantile(x, 0.025), 
    `97.5%` = quantile(x, 0.975))
}

# 3. Apply the function to the lp__ parameter across the chains (dimension 2)
cat("\n--- lp__ Summary Per Chain ---\n")
lp_summary <- apply(post_array[, , "lp__"], MARGIN = 2, FUN = summarize_chain)
print(round(lp_summary, 2))

# 4. Do the same for rho to see exactly how Chain 4 differs numerically
cat("\n--- rho Summary Per Chain ---\n")
rho_summary <- apply(post_array[, , "rho"], MARGIN = 2, FUN = summarize_chain)
print(round(rho_summary, 3))
```

Clearly significantly lower for the 4th chain on the lp, suggesting it has reached a local minimum.

Below is code to show density for rho for the different chains, along with some more diagnostic plots of the log posterior.

```{r}

# Load the required libraries
library(bayesplot)
library(ggplot2)

# 1. Convert your Stan fit object to an array (bayesplot prefers this format)
# Replace 'out_joint' with whatever you named your loaded .rds model
posterior_array <- as.array(out_lv)

# 2. Plot the overlaid densities for each chain
mcmc_dens_overlay(posterior_array, pars = c("rho")) +
  ggtitle("Posterior Density of rho (Shrinkage Factor) per Chain") +
  theme_minimal()

# Optional: If you also want to see the combined density of just the "good" chains
# (assuming chains 2, 3, and 4 are the converged ones)
mcmc_dens(posterior_array[, 2:4, ], pars = c("rho")) +
  ggtitle("Combined Posterior Density of rho (Chains 2-4 only)") +
  theme_minimal()





# Assuming your array is loaded as 'posterior_array'
# (e.g., posterior_array <- as.array(out_joint))

# Plot 1: The Elevation Trace
# Are all chains at the same height?
mcmc_trace(posterior_array, pars = "lp__") +
  ggtitle("Traceplot of Log-Posterior (lp__) per Chain") +
  theme_minimal()

# Plot 2: The Elevation Density
# Do the distributions perfectly overlap, or is one shifted left (lower)?
mcmc_dens_overlay(posterior_array, pars = "lp__") +
  ggtitle("Density of Log-Posterior (lp__) per Chain") +
  theme_minimal()

# Plot 3: The Scatterplot (Bob's suggestion)
# This maps the GPS coordinate (rho) against the Elevation (lp__)
# If you see two distinct clouds, you have a multimodal posterior!
mcmc_scatter(posterior_array, pars = c("rho", "lp__")) +
  ggtitle("Geometry of the Posterior: rho vs lp__") +
  theme_minimal()
```

Now I'll check the results for the first model that didn't work for 2 latent variables on both morphology and fitness. This is from the R script combined_test_model.R (or that script might actually contain the 1 lv version which i did afterwards...). What I need to check is the chains and also the rho densities and also the log posteriors, and maybe more.

```{r}
out_old <- readRDS('output_markov/Output_Joint_Starter.rds')

print(out_old,
      # pars = c(               "gamma_repro",       # Selection gradient: Size -> Reproduction!
      #          "gamma_surv",        # Selection gradient: Size -> Survival!
      #          "Lambda",            # Factor loadings (How traits build Size)
      #          "h2_lv",             # Heritability of the Latent Size
      #          "rho",               # Shrinkage factor
      #          "sd_R",              # Residual error for morph traits
      #          "sd_year_morph",     # Year variance for morphology
      #          "sd_year_repro",     # Year variance for reproduction
      #          "sd_year_surv",      # Year variance for survival
      #          "sd_init_morph",     # Measurer variance for morphology
      #          "lp__"),             # Log-posterior (model fit)
      probs = c(0.025, 0.975),
      digits_summary = 3)
```

```{r}
library(bayesplot)
library(ggplot2)

# 1. Plot a stable parameter (The Grand Mean)
color_scheme_set("mix-blue-red")
p1 <- mcmc_trace(out_old, pars = c("alpha_trait[1]", "sd_year_surv")) +
  ggtitle("Stable Parameters (Rhat = 1.00)")

# 2. Plot the 'Flipped' parameters
p2 <- mcmc_trace(out_old, pars = c("Lambda[1,1]", "gamma_repro[1]", "gamma_surv[1]")) +
  ggtitle("Sign-Flipped Parameters (High Rhat)")

p3 <- mcmc_trace(out_old, pars = c("Lambda[1,2]", "Lambda[3,2]", "gamma_repro[2]", "gamma_surv[2]"))

p4 <- mcmc_trace(out_old, pars = c("Lambda[2,1]", "Lambda[1,2]", "Lambda[3,1]", "Lambda[2,2]"))

  

# Show the plots
print(p1)
print(p2)
print(p3)
print(p4)
```

```{r}

# Load the required libraries
library(bayesplot)
library(ggplot2)

# 1. Convert your Stan fit object to an array (bayesplot prefers this format)
# Replace 'out_joint' with whatever you named your loaded .rds model
posterior_array <- as.array(out_old)

# 2. Plot the overlaid densities for each chain
mcmc_dens_overlay(posterior_array, pars = c("rho")) +
  ggtitle("Posterior Density of rho (Shrinkage Factor) per Chain") +
  theme_minimal()

# Optional: If you also want to see the combined density of just the "good" chains
# (assuming chains 2, 3, and 4 are the converged ones)
mcmc_dens(posterior_array[, 2:4, ], pars = c("rho")) +
  ggtitle("Combined Posterior Density of rho (Chains 2-4 only)") +
  theme_minimal()





# Assuming your array is loaded as 'posterior_array'
# (e.g., posterior_array <- as.array(out_joint))

# Plot 1: The Elevation Trace
# Are all chains at the same height?
mcmc_trace(posterior_array, pars = "lp__") +
  ggtitle("Traceplot of Log-Posterior (lp__) per Chain") +
  theme_minimal()

# Plot 2: The Elevation Density
# Do the distributions perfectly overlap, or is one shifted left (lower)?
mcmc_dens_overlay(posterior_array, pars = "lp__") +
  ggtitle("Density of Log-Posterior (lp__) per Chain") +
  theme_minimal()

# Plot 3: The Scatterplot (Bob's suggestion)
# This maps the GPS coordinate (rho) against the Elevation (lp__)
# If you see two distinct clouds, you have a multimodal posterior!
mcmc_scatter(posterior_array, pars = c("rho", "lp__")) +
  ggtitle("Geometry of the Posterior: rho vs lp__") +
  theme_minimal()


# 1. Extract the 3D array from your Stan fit

# 2. Define a quick function to get the Mean and 95% CI
summarize_chain <- function(x) {
  c(Mean = mean(x), 
    `2.5%` = quantile(x, 0.025), 
    `97.5%` = quantile(x, 0.975))
}

# 3. Apply the function to the lp__ parameter across the chains (dimension 2)
cat("\n--- lp__ Summary Per Chain ---\n")
lp_summary <- apply(posterior_array[, , "lp__"], MARGIN = 2, FUN = summarize_chain)
print(round(lp_summary, 2))

# 4. Do the same for rho to see exactly how Chain 4 differs numerically
cat("\n--- rho Summary Per Chain ---\n")
rho_summary <- apply(posterior_array[, , "rho"], MARGIN = 2, FUN = summarize_chain)
print(round(rho_summary, 3))

```

So, it looks like, contrary to the latest run, this older run actually had two equally good modes, as you can see that the posterior mean for the lp for chain 1 (doing its own thing) is almost exactly the same as the others.

Now I'll run the 2 LV model with survival and reproduction but without any fixed effects, just measurer initials random effect and year random effect.

```{stan}
data {
  // Dimensions
  int<lower=1> No_morph; 
  int<lower=1> No_repro;
  int<lower=1> No_surv;
  int<lower=1> Nt;  
  int<lower=1> Nlv; 
  int<lower=1> Na;  

  int<lower=0> K_morph;
  int<lower=0> K_repro;
  int<lower=0> K_surv;
  
  matrix[No_morph, K_morph] X_morph;
  matrix[No_repro, K_repro] X_repro;
  matrix[No_surv, K_surv] X_surv;
  
  // Responses
  matrix[No_morph, Nt] Y_morph; 
  array[No_repro] int<lower=0> Y_repro; // Poisson count
  array[No_surv] int<lower=0, upper=1> Y_surv; // Bernoulli 0/1
  
  // Pedigree & Links
  array[No_morph] int<lower=1, upper=Na> animal_morph; 
  array[No_repro] int<lower=1, upper=Na> animal_repro;
  array[No_surv]  int<lower=1, upper=Na> animal_surv;
  
  array[Na] int<lower=1, upper=Na+1> dam; 
  array[Na] int<lower=1, upper=Na+1> sire;
  vector[Na] dii; 
  
  // Random Effects Indices
  int<lower=1> N_year;
  array[No_morph] int<lower=1, upper=N_year> year_morph;
  array[No_repro] int<lower=1, upper=N_year> year_repro;
  array[No_surv]  int<lower=1, upper=N_year> year_surv;
  
  int<lower=1> N_init;
  array[No_morph] int<lower=1, upper=N_init> init_morph;
}

parameters {
  // Intercepts (Since we dropped the X matrix)
  
  matrix[K_morph, Nt] beta_morph;
  vector[K_repro] beta_repro;
  vector[K_surv] beta_surv;

  // Selection Gradients (The link between Size and Fitness)
  vector[Nlv] gamma_repro;
  vector[Nlv] gamma_surv;

  // Variances
  vector<lower=0>[Nt] sd_R; 
  real<lower=0, upper=1> rho;
  vector<lower=0, upper=1>[Nlv] h2_lv; 
  
  matrix[Nt, Nlv] lambda_raw;
  
  real<lower=0> lambda_1_1;
  real<lower=0> lambda_2_2;

  // Latent Effects
  matrix[Nlv, Na] w_a;  
  matrix[Nlv, Na] w_pe; 
  
  // Random Effects (Year and Measurer)
  matrix[N_year, Nt] z_year_morph;
  vector<lower=0>[Nt] sd_year_morph;
  
  vector[N_year] z_year_repro;
  real<lower=0> sd_year_repro;
  
  vector[N_year] z_year_surv;
  real<lower=0> sd_year_surv;
  
  matrix[N_init, Nt] z_init_morph;
  vector<lower=0>[Nt] sd_init_morph;
}

transformed parameters {
  // 1. Pedigree Recursion
  matrix[Na, Nlv] G_effect; 
  {
    matrix[Na+1, Nlv] aug_a;
    aug_a[Na+1] = rep_row_vector(0, Nlv);
    for (i in 1:Na) {
       aug_a[i] = 0.5 * aug_a[dam[i]] + 0.5 * aug_a[sire[i]] 
                  + sqrt(dii[i]) * w_a[, i]'; 
    }
    G_effect = aug_a[1:Na];
  }

  // 2. Latent Factor
  matrix[Na, Nlv] LV;
  for(k in 1:Nlv) {
    LV[, k] = sqrt(h2_lv[k]) * G_effect[, k] + sqrt(1 - h2_lv[k]) * w_pe[k, ]';
  }

  matrix[Nt, Nlv] lambda_con;
  lambda_con = lambda_raw;
  lambda_con[1,1] = lambda_1_1;
  lambda_con[2,2] = lambda_2_2;
  
  // 3. Shrinkage
  matrix[Nt, Nlv] Lambda;
  Lambda[, 1] = lambda_con[, 1];            
  Lambda[, 2] = lambda_con[, 2] * sqrt(rho);
}

model {
  // Priors
  rho ~ beta(1, 2);
  h2_lv ~ beta(2.5, 2.5); 
  to_vector(lambda_raw) ~ std_normal();
  sd_R ~ normal(0, 1);
  
  to_vector(beta_morph) ~ normal(0,1);
  beta_repro ~ normal(0,1);
  beta_surv ~ normal(0,1);
  
  gamma_repro ~ normal(0, 1);
  gamma_surv ~ normal(0, 1);

  to_vector(w_a) ~ std_normal();
  to_vector(w_pe) ~ std_normal();
  to_vector(z_year_morph) ~ std_normal();
  z_year_repro ~ std_normal();
  z_year_surv ~ std_normal();
  to_vector(z_init_morph) ~ std_normal();
  
  sd_year_morph ~ exponential(2); 
  sd_year_repro ~ exponential(2); 
  sd_year_surv ~ exponential(2); 
  sd_init_morph ~ exponential(2);

  // --- LIKELIHOODS ---
  
  // 1. Morphology
  for(i in 1:No_morph){
    for(t in 1:Nt){
      real mu =  
                dot_product(X_morph[i], beta_morph[, t])
                + LV[animal_morph[i], 1] * Lambda[t, 1] 
                + LV[animal_morph[i], 2] * Lambda[t, 2]
                + z_year_morph[year_morph[i], t] * sd_year_morph[t]
                + z_init_morph[init_morph[i], t] * sd_init_morph[t];
      Y_morph[i, t] ~ normal(mu, sd_R[t]);
    }
  }
  
// // 2. Reproduction (Poisson)
  for(i in 1:No_repro){
    real log_lambda =  dot_product(X_repro[i], beta_repro)
                      + gamma_repro[1] * LV[animal_repro[i], 1]
                      + gamma_repro[2] * LV[animal_repro[i], 2]
                      + z_year_repro[year_repro[i]] * sd_year_repro;
    Y_repro[i] ~ poisson_log(log_lambda);
  }

  // 3. Survival (Bernoulli)
  for(i in 1:No_surv){
    real logit_p = dot_product(X_surv[i], beta_surv)
                    + gamma_surv[1] * LV[animal_surv[i], 1]
                   + gamma_surv[2] * LV[animal_surv[i], 2]
                   + z_year_surv[year_surv[i]] * sd_year_surv;
    Y_surv[i] ~ bernoulli_logit(logit_p);
  }

}
```

Here are the results from this latest run. This is a run where I put a beta(1,2) prior on rho. And I didn't include that many fixed effects (or really any). The script is the one called something with "hopefully".

```{r}
out_hopeful <- readRDS("output_markov/Output_hopefully.rds")
print(out_hopeful, probs = c(0.025, 0.975), digits_summary = 3)
```

Now, this looks like some issues with mixing again, so we need to investigate. It seems the betas and the residuals are looking fine though, the and rho is also looking okay, so it might just be a case of sign flipping (although that shouldn't be possible anymore). So we need to look at the chains.

```{r}
library(bayesplot)
library(ggplot2)

color_scheme_set("mix-blue-red")

chains1 <- mcmc_trace(out_hopeful, pars = c("Lambda[1,1]", "Lambda[1,2]", "Lambda[3,1]", "Lambda[3,2]"))
chains2 <- mcmc_trace(out_hopeful, pars = c("gamma_repro[1]", "gamma_repro[2]", "gamma_surv[1]", "gamma_surv[2]", "rho"))
chains3 <- mcmc_trace(out_hopeful, pars = c("Lambda[2,1]", "Lambda[2,2]", "Lambda[4,1]", "Lambda[4,2]", "Lambda[5,1]", "Lambda[5,2]"))
chains4 <- mcmc_trace(out_hopeful, pars= c("lp__", "h2_lv[1]", "h2_lv[2]"))

print(chains1)
print(chains2)
print(chains3)
print(chains4)



```

Alright, it looks like there are two possible modes that have multiple chains each. So we need to check some densities and log posteriors.

```{r}
posterior_array_2 <- as.array(out_hopeful)

mcmc_dens_overlay(posterior_array_2, pars = c("rho")) +
  ggtitle("Posterior Density of rho (Shrinkage Factor) per Chain") +
  theme_minimal()

```

Alright it looks like there are 3 chains in each case, and they seem to slightly disagree on the distribution of rho.

```{r}
mcmc_dens_overlay(posterior_array_2, pars = "lp__") +
  ggtitle("Density of Log-Posterior (lp__) per Chain") +
  theme_minimal()
```

Okay, so the log posteriors seem to be slightly different for the two modes.

```{r}
summarize_chain <- function(x) {
  c(Mean = mean(x), 
    `2.5%` = quantile(x, 0.025), 
    `97.5%` = quantile(x, 0.975))
}

# 3. Apply the function to the lp__ parameter across the chains (dimension 2)
cat("\n--- lp__ Summary Per Chain ---\n")
lp_summary <- apply(posterior_array_2[, , "lp__"], MARGIN = 2, FUN = summarize_chain)
print(round(lp_summary, 2))

# 4. Do the same for rho to see exactly how Chain 4 differs numerically
cat("\n--- rho Summary Per Chain ---\n")
rho_summary <- apply(posterior_array_2[, , "rho"], MARGIN = 2, FUN = summarize_chain)
print(round(rho_summary, 3))
```

It seems the mode with rho around 0.28 has somewhat higher log posteriors, by approximately 5 log units, compared to the mode with rho around 0.34-0.35, and 5 log units is a significant difference. But the differences within each mode is also quite large, so the 5 log unit difference is quite fuzzy.

Now, one question is, can I just take the log posterior and say that, hey, this mode has a significantly higher log posterior, so this is the better mode, and then ignore the other one? If not, how do I prevent this duality in the run? Should I just tweak the priors even more or the initialization to avoid the model from going into the "bad" mode?

But in any case, it seems the model finds evidence for a second latent variable for these data, although that isn't exactly shown without a regularization or something, I suppose.

What is interesting is that even though the model seems to respect the rho, the loadings seem to be about the same magnitude in both modes. If the hypothesis that the body size is the latent variable representing the most variance, it would not make sense that the rho would be only about 0.3 if the first LV represents shape in one of the modes. Then the rho would have to be like 3 or something. One possibility is that it is the lambda_raw that gets blown up, to accommodate the requirement for both rho and the actual loadings. But the lambda_raw has a standard normal prior (I think), this might need to be tightened, or I'm not exactly sure what can be done except constraining some loadings.

Alright, one possibility is to constrain the priors of the loadings for the second latent variable to have a lower standard deviation, and forget about rho (or estimate it after the fact could be done probably). Another option could potentially be to constrain lambda_raw (or equivalent) to only be sampled within a smaller interval of say -1.5 to 1.5.

By the way, I think it actually was the model with the body size as the second latent variable that had the highest log posterior in the latest run, but need to doublecheck this.

*Run with just stricter priors on the second factor*

This is how the model was done.

```{stan}
parameters {
  // Intercepts (Since we dropped the X matrix)
  
  matrix[K_morph, Nt] beta_morph;
  vector[K_repro] beta_repro;
  vector[K_surv] beta_surv;

  // Selection Gradients (The link between Size and Fitness)
  vector[Nlv] gamma_repro;
  vector[Nlv] gamma_surv;

  // Variances
  vector<lower=0>[Nt] sd_R; 
  // real<lower=0, upper=1> rho;
  vector<lower=0, upper=1>[Nlv] h2_lv; 
  
  matrix[Nt, Nlv] lambda_raw;
  
  real<lower=0> lambda_1_1;
  vector[Nt-1] lambda_free_1;
  real<lower=0> lambda_2_2;
  vector[Nt-1] lambda_free_2;


  // Latent Effects
  matrix[Nlv, Na] w_a;  
  matrix[Nlv, Na] w_pe; 
  
  // Random Effects (Year and Measurer)
  matrix[N_year, Nt] z_year_morph;
  vector<lower=0>[Nt] sd_year_morph;
  
  vector[N_year] z_year_repro;
  real<lower=0> sd_year_repro;
  
  vector[N_year] z_year_surv;
  real<lower=0> sd_year_surv;
  
  matrix[N_init, Nt] z_init_morph;
  vector<lower=0>[Nt] sd_init_morph;
}

transformed parameters {
  // 1. Pedigree Recursion
  matrix[Na, Nlv] G_effect; 
  {
    matrix[Na+1, Nlv] aug_a;
    aug_a[Na+1] = rep_row_vector(0, Nlv);
    for (i in 1:Na) {
       aug_a[i] = 0.5 * aug_a[dam[i]] + 0.5 * aug_a[sire[i]] 
                  + sqrt(dii[i]) * w_a[, i]'; 
    }
    G_effect = aug_a[1:Na];
  }

  // 2. Latent Factor
  matrix[Na, Nlv] LV;
  for(k in 1:Nlv) {
    LV[, k] = sqrt(h2_lv[k]) * G_effect[, k] + sqrt(1 - h2_lv[k]) * w_pe[k, ]';
  }


  matrix[Nt, Nlv] Lambda = rep_matrix(0.0, Nt, Nlv);
  
  Lambda[1,1] = lambda_1_1;
  
  for (t in 2:Nt) {
    Lambda[t, 1] = lambda_free_1[t-1];
  }
  
  Lambda[2,2] = lambda_2_2;
  Lambda[1,2] = lambda_free_2[1];
  
  for (t in 3:Nt) {
    Lambda[t, 2] = lambda_free_2[t-2];
  }

}

model {
  // Priors
  // rho ~ beta(1, 2);
  h2_lv ~ beta(2.5, 2.5); 
  lambda_1_1 ~ normal(0, 1.0);
  lambda_free_1 ~ normal(0, 1.0);
  
  lambda_2_2 ~ normal(0, 0.2);
  lambda_free_2 ~ normal(0, 0.2);

  sd_R ~ normal(0, 1);
  
  to_vector(beta_morph) ~ normal(0,1);
  beta_repro ~ normal(0,1);
  beta_surv ~ normal(0,1);
  
  gamma_repro ~ normal(0, 1);
  gamma_surv ~ normal(0, 1);

  to_vector(w_a) ~ std_normal();
  to_vector(w_pe) ~ std_normal();
  to_vector(z_year_morph) ~ std_normal();
  z_year_repro ~ std_normal();
  z_year_surv ~ std_normal();
  to_vector(z_init_morph) ~ std_normal();
  
  sd_year_morph ~ exponential(2); 
  sd_year_repro ~ exponential(2); 
  sd_year_surv ~ exponential(2); 
  sd_init_morph ~ exponential(2);
  
  
  // --- LIKELIHOODS ---
  
  // 1. Morphology
  for(i in 1:No_morph){
    for(t in 1:Nt){
      real mu =  
                dot_product(X_morph[i], beta_morph[, t])
                + LV[animal_morph[i], 1] * Lambda[t, 1] 
                + LV[animal_morph[i], 2] * Lambda[t, 2]
                + z_year_morph[year_morph[i], t] * sd_year_morph[t]
                + z_init_morph[init_morph[i], t] * sd_init_morph[t];
      Y_morph[i, t] ~ normal(mu, sd_R[t]);
    }
  }
  
// // 2. Reproduction (Poisson)
  for(i in 1:No_repro){
    real log_lambda =  dot_product(X_repro[i], beta_repro)
                      + gamma_repro[1] * LV[animal_repro[i], 1]
                      + gamma_repro[2] * LV[animal_repro[i], 2]
                      + z_year_repro[year_repro[i]] * sd_year_repro;
    Y_repro[i] ~ poisson_log(log_lambda);
  }

  // 3. Survival (Bernoulli)
  for(i in 1:No_surv){
    real logit_p = dot_product(X_surv[i], beta_surv)
                    + gamma_surv[1] * LV[animal_surv[i], 1]
                   + gamma_surv[2] * LV[animal_surv[i], 2]
                   + z_year_surv[year_surv[i]] * sd_year_surv;
    Y_surv[i] ~ bernoulli_logit(logit_p);
  }

}
```

Alright, now I ran a model without rho (just computing it after the fact as a generated quantity), and just using much narrower priors on the loadings of the second latent variable.

```{r}
out_just_prior <- readRDS("output_markov/Output_hopefully_no_rho.rds")

print(out_just_prior, probs = c(0.025, 0.975), digits_summary = 3)
```

However, I had constrained the loadings lambda_1_1 and lambda_2_2 to be positive, which caused the model to be forced to have the second factor be the body size factor, and inflated the rho to around 1.9 even though I had a normal(0, 0.2) prior on the loadings of the second factor and a standard normal on the loadings of the first factor, causing the model to be forced to fight the prior. So it seems the best way after all is to just fix any sign flipping after the run, and not constrain any loadings to be positive (or zero).

Now, the results show that the factor that was the shape factor actually had a positive correlation with the wing and negative with the others, suggesting a different shape factor than the ones previously found (I think). Whether this means that there are grounds for a third latent variable or whether this was a result of the weird model fitting caused by the positive constraint is not clear to me, but it seems it is a valid factor in any case. It could also suggest that the ground for a single shape factor is weak, and when forced to find a single shape factor, the model jumps between different possibilities.

The next step then, is to run the model without any positive constraints, but I'll ask Bob first.

When it comes to why the model with the rho using a beta prior had difficulties settling between the two modes, I think the best explanation is that it could just inflate the lambda_raw, to accommodate the requirement for both rho and the actual loadings, and that the model had difficulties finding a good balance between these two, which is why it had difficulties settling on one mode.

Okay, now I will change it so that I don't have the positive constraints on any lambdas.

Alright, now here are the results of the run without any positive constraints.

```{r}

out_no_pos <- readRDS("output_markov/Output_hopefully_no_pos.rds")


print(out_no_pos, probs = c(0.025, 0.975), digits_summary = 3)
```

Seems like there could be a sign flipping issue (or hopefully that's what it is).

Now, I'll check the chains.

```{r}

library(bayesplot)
library(ggplot2)

color_scheme_set("mix-blue-red")

chains1 <- mcmc_trace(out_no_pos, pars = c("Lambda[1,1]", "Lambda[1,2]", "Lambda[3,1]", "Lambda[3,2]"))

chains2 <- mcmc_trace(out_no_pos, pars = c("gamma_repro[1]", "gamma_repro[2]", "gamma_surv[1]", "gamma_surv[2]", "rho"))

chains3 <- mcmc_trace(out_no_pos, pars = c("Lambda[2,1]", "Lambda[2,2]", "Lambda[4,1]", "Lambda[4,2]", "Lambda[5,1]", "Lambda[5,2]"))




chains4 <- mcmc_trace(out_no_pos, pars= c("lp__", "h2_lv[1]", "h2_lv[2]"))

print(chains1)
print(chains2)
print(chains3)
print(chains4)

```

So it is clear from these results that the second latent variable is sign flipped between the chains. Now, I will correct the chains that are flipped for the second variable. The flipped chain are 2 and 4 vs 1 and 3. So I will flip the sign of the second latent variable for chains 2 and 4, and then combine the chains and check the results.

And as can be seen from the log posteriors, the chains all have around the same value, so it doesn't seem like there are multiple different universes that are being explored.

```{r}

# Extract the posterior samples as an array
posterior_array_no_pos <- as.array(out_no_pos)
# Flip the sign of the second latent variable for chains 2 and 4
posterior_array_no_pos[, c(2, 4), "Lambda[1,2]"] <- -posterior_array_no_pos[, c(2, 4), "Lambda[1,2]"]
posterior_array_no_pos[, c(2, 4), "Lambda[2,2]"] <- -posterior_array_no_pos[, c(2, 4), "Lambda[2,2]"]
posterior_array_no_pos[, c(2, 4), "Lambda[3,2]"] <- -posterior_array_no_pos[, c(2, 4), "Lambda[3,2]"]
posterior_array_no_pos[, c(2, 4), "Lambda[4,2]"] <- -posterior_array_no_pos[, c(2, 4), "Lambda[4,2]"]
posterior_array_no_pos[, c(2, 4), "Lambda[5,2]"] <- -posterior_array_no_pos[, c(2, 4), "Lambda[5,2]"]

posterior_array_no_pos[, c(2, 4), "gamma_repro[2]"] <- -posterior_array_no_pos[, c(2, 4), "gamma_repro[2]"]
posterior_array_no_pos[, c(2, 4), "gamma_surv[2]"] <- -posterior_array_no_pos[, c(2, 4), "gamma_surv[2]"]

```

Then we have to find the results with the 2 chains flipped.

```{r}
# 1. Combine the corrected array into a summary table
# We use rstan::monitor to re-calculate R-hat and n_eff for the modified parameters
# (Note: we only modified Lambda and gamma, so the others will stay the same)
library(rstan)

# Convert the modified array to a monitor object
# 'warmup = 0' because we are using the post-warmup array already
summary_corrected <- monitor(posterior_array_no_pos, warmup = 0, digits_summary = 3)

# 2. View the corrected results for the parameters you flipped
# This should now show R-hats very close to 1.000
cat("\n--- Corrected Results for Lambda (Factor 2) and Gammas ---\n")
params_to_check <- c("Lambda[1,2]", "Lambda[2,2]", "Lambda[3,2]", 
                     "Lambda[4,2]", "Lambda[5,2]", 
                     "gamma_repro[2]", "gamma_surv[2]")



```

```{r}
library(posterior)

# 1. Convert to draws array
draws_corrected <- as_draws_array(posterior_array_no_pos)

# 2. Get the summary
summary_tab <- summarise_draws(draws_corrected)

# 3. THE FIX: Strip the fancy 'posterior' formatting and make it a basic data frame
summary_df <- as.data.frame(summary_tab)

# 4. Subset and print safely
params_to_check <- c("Lambda[1,2]", "Lambda[2,2]", "Lambda[3,2]", 
                     "Lambda[4,2]", "Lambda[5,2]", 
                     "gamma_repro[2]", "gamma_surv[2]")

subset_summary <- summary_df[summary_df$variable %in% params_to_check, ]
print(subset_summary)
```

```{r}
library(bayesplot)
library(ggplot2)

# Set a color scheme so it's easy to distinguish the 4 chains
color_scheme_set("mix-blue-red")

# Select the variables with the worst R-hats and lowest ESS
worst_vars <- c("Lambda[1,2]", "Lambda[2,2]", "Lambda[4,2]", "gamma_surv[2]")

# Plot the modified array
plot_flipped_chains <- mcmc_trace(posterior_array_no_pos, pars = worst_vars) +
  ggtitle("Traceplots AFTER Sign Correction (Chains 2 & 4 Flipped)") +
  theme_minimal()

print(plot_flipped_chains)
```

It looks like we're getting slow mixing and high autocorrelation. It doesn't seem like any of the chains are living in an entirely different universe from the others, as the chains are overlapping (no clear gap between them). We are getting R hats as high as 1.15, and n effs as low as 27, so it's not ideal. But running for more iterations could potentially help.

Here is a function to hopefully do the post-run sign flipping automatically for the chains that are flipped:

```{r}
auto_align_signs <- function(stan_array, factor_num, anchor_trait, n_traits = 5) {
  # Get the number of chains from the array dimensions
  n_chains <- dim(stan_array)[2]
  
  # Define the anchor parameter we are checking (e.g., "Lambda[1,1]")
  anchor_param <- paste0("Lambda[", anchor_trait, ",", factor_num, "]")
  
  for (c in 1:n_chains) {
    # Check if the mean of the anchor loading in this chain is negative
    if (mean(stan_array[, c, anchor_param]) < 0) {
      message(sprintf("Aligning Chain %d: Flipping signs for Factor %d", c, factor_num))
      
      # 1. Flip all Loadings (Lambda) for this factor
      for (t in 1:n_traits) {
        lam_name <- paste0("Lambda[", t, ",", factor_num, "]")
        stan_array[, c, lam_name] <- -stan_array[, c, lam_name]
      }
      
      # 2. Flip the Selection Gradients (gamma) for this factor
      gamma_repro_name <- paste0("gamma_repro[", factor_num, "]")
      gamma_surv_name  <- paste0("gamma_surv[", factor_num, "]")
      
      # We use if() checks just in case a model doesn't have both gammas
      if (gamma_repro_name %in% dimnames(stan_array)$parameters) {
        stan_array[, c, gamma_repro_name] <- -stan_array[, c, gamma_repro_name]
      }
      if (gamma_surv_name %in% dimnames(stan_array)$parameters) {
        stan_array[, c, gamma_surv_name] <- -stan_array[, c, gamma_surv_name]
      }
    }
  }
  
  return(stan_array)
}
```

Now it's time to look at the most recent results, in which we ran for 20,000 total iterations in order to check if that helps the convergence.

```{r}
out_long <- readRDS("output_markov/Output_hopefully_no_pos_long.rds")
print(out_long, probs = c(0.025, 0.975), digits_summary = 3)
```

Alright, we have high rhat's as usual, now let's take a look at the chains.

```{r}
library(bayesplot)
library(ggplot2)
color_scheme_set("mix-blue-red")
chains1_long <- mcmc_trace(out_long, pars = c("Lambda[1,1]", "Lambda[1,2]", "Lambda[3,1]", "Lambda[3,2]"))
chains2_long <- mcmc_trace(out_long, pars = c("gamma_repro[1]", "gamma_repro[2]", "gamma_surv[1]", "gamma_surv[2]", "rho"))
chains3_long <- mcmc_trace(out_long, pars = c("Lambda[2,1]", "Lambda[2,2]", "Lambda[4,1]", "Lambda[4,2]", "Lambda[5,1]", "Lambda[5,2]"))
chains4_long <- mcmc_trace(out_long, pars= c("lp__", "h2_lv[1]", "h2_lv[2]"))
print(chains1_long)
print(chains2_long)
print(chains3_long)
print(chains4_long)
```

It looks mostly good, but there is some sign flipping. But what is interesting is that chain 4 took a lot longer than the 3 others, but it looks like it's doing mostly the same thing as the others, although the gammas on the second latent variable might be acting differently for chain 4, but it's kind of hard to tell. In any case, let's check the sign flipped version.

```{r}

library(posterior)

# 1. Convert the raw Stan output to a 3D array (iterations x chains x parameters)
raw_array <- as.array(out_long)

# 2. Align Factor 1 (The "Size/Shape" Factor)
# We use Trait 1 (Wing Length) as the anchor. If a chain made Wing Length negative, 
# this flips that entire chain for Factor 1.
aligned_array <- auto_align_signs(stan_array = raw_array, 
                                  factor_num = 1, 
                                  anchor_trait = 1, 
                                  n_traits = 5)

# 3. Align Factor 2 (The second Factor)
# We use Trait 2 (Tarsus) as the anchor here, since Wing Length might have 
# a negative or near-zero loading on the shape factor depending on the rotation.
aligned_array <- auto_align_signs(stan_array = aligned_array, 
                                  factor_num = 2, 
                                  anchor_trait = 2, 
                                  n_traits = 5)

# 4. Convert the aligned array into a clean 'posterior' draws object
clean_draws <- as_draws_array(aligned_array)

# 5. Generate the summary table (forcing it to a standard data frame to avoid R errors)
summary_df <- as.data.frame(summarise_draws(clean_draws))

# 6. Print the critical parameters to verify convergence
params_to_check <- c("Lambda[1,1]", "Lambda[1,2]", "Lambda[2,2]", "Lambda[3,2]", "Lambda[4,2]", "Lambda[5,2]", "Lambda[2,1]", "Lambda[3,1]", "Lambda[4,1]", "Lambda[5,1]",
                     "gamma_repro[1]", "gamma_repro[2]", 
                     "gamma_surv[1]", "gamma_surv[2]", "rho", "lp__", "h2_lv[1]", "h2_lv[2]", 
                     "sd_R[1]", "sd_R[2]", "sd_R[3]", "sd_R[4]", "sd_R[5]")

print(summary_df[summary_df$variable %in% params_to_check, 
                 c("variable", "mean", "rhat", "ess_bulk")])
```

```{r}
library(bayesplot)
library(posterior)
library(ggplot2)

# Assuming 'clean_draws' is your aligned array from the previous step
color_scheme_set("mix-blue-red")

# 1. Traceplots to visually confirm the alignment worked perfectly
pars_to_plot <- c("Lambda[1,2]", "Lambda[2,2]", "gamma_surv[1]", "gamma_surv[2]")
p_traces <- mcmc_trace(clean_draws, pars = pars_to_plot) +
  ggtitle("Perfectly Aligned Traceplots (Post-HPC)") +
  theme_minimal()
print(p_traces)

# 2. Extract specific Credible Intervals (2.5% and 97.5%)
# We define a custom function for summarise_draws to get the exact quantiles
summary_ci <- summarise_draws(
  clean_draws, 
  "mean", 
  ~quantile(.x, probs = c(0.025, 0.975)),
  "rhat", 
  "ess_bulk"
)

```

```{r}
# Turn off scientific notation globally
options(scipen = 999)

# Convert to data frame to prevent printing errors
summary_ci_df <- as.data.frame(summary_ci)

# Filter to your key parameters
params_to_check <- c(
  "Lambda[1,1]", "Lambda[1,2]", "Lambda[2,2]", "Lambda[3,2]", "Lambda[4,2]", "Lambda[5,2]", "Lambda[2,1]", "Lambda[3,1]", "Lambda[4,1]", "Lambda[5,1]",
                     "gamma_repro[1]", "gamma_repro[2]", 
                     "gamma_surv[1]", "gamma_surv[2]", "rho", "lp__", "h2_lv[1]", "h2_lv[2]", 
                     "sd_R[1]", "sd_R[2]", "sd_R[3]", "sd_R[4]", "sd_R[5]"
)

subset_ci <- summary_ci_df[summary_ci_df$variable %in% params_to_check, ]

# Find all numeric columns and round them to 3 decimal places
is_num <- sapply(subset_ci, is.numeric)
subset_ci[is_num] <- lapply(subset_ci[is_num], round, 3)

# Print the final, publication-ready table
print(subset_ci, row.names = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)

# 1. Filter the 'subset_ci' dataframe for just the selection gradients
gamma_df <- subset_ci %>%
  filter(grepl("^gamma", variable)) %>%
  mutate(
    # Create clean, readable labels for the y-axis
    Trait = case_when(
      variable == "gamma_surv[1]" ~ "Survival: Factor 1 (Size)",
      variable == "gamma_surv[2]" ~ "Survival: Factor 2 (Shape)",
      variable == "gamma_repro[1]" ~ "Reproduction: Factor 1 (Size)",
      variable == "gamma_repro[2]" ~ "Reproduction: Factor 2 (Shape)"
    ),
    # Create a grouping variable so we can color-code Survival vs Reproduction
    Fitness_Type = ifelse(grepl("surv", variable), "Survival", "Reproduction")
  )

# 2. Build the ggplot
forest_plot <- ggplot(gamma_df, aes(x = mean, y = Trait, color = Fitness_Type)) +
  # Draw the critical 'Zero' line first so it sits behind the data
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgray", linewidth = 1) +
  
  # Draw the 95% Credible Intervals using backticks for the percentage column names
  geom_errorbarh(aes(xmin = `2.5%`, xmax = `97.5%`), height = 0.2, linewidth = 1.2) +
  
  # Draw the Mean estimates (the dots)
  geom_point(size = 4) +
  
  # Custom colors (Colorblind-friendly blue and orange)
  scale_color_manual(values = c("Survival" = "#0072B2", "Reproduction" = "#D55E00")) +
  
  # Clean up the labels
  labs(
    title = "Selection on Latent Size and Shape",
    x = "Selection Gradient (Estimate \u00B1 95% CI)",
    y = NULL
  ) +
  
  # Apply a clean, minimalist theme suitable for a thesis
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none", # Hide legend since the y-axis labels explain it
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", color = "black"),
    axis.text.x = element_text(color = "black")
  )

# Print the plot!
print(forest_plot)
```


