---
title: "Shrink Prior"
author: "Elias Ovesen"
editor: visual
---

Now, I'm running the model with the shrinkage prior (gamma priors). The Stan script used is the one called "gamma_shrink.stan", and the R script I used was the one called "gamma_shrink_script.R". I would ideally try to 1) send only the relevant data to the HPC (Markov), and also only put the relevant code changes here and import the rds data set, but still not comfortable with that workflow. I have to run it on the HPC anyway, so should I run any of the model iterations/variations here, or only look at the data? I in any case should separate the code into separate modular chunks so I don't have to prepare the data, pedigree etc. every single time. I guess the actual model code shouldn't be here, as this is where I investigate the results, and analyze and write about them and the theory around what I'm doing.

But actually developing the model also requires theory and analysis, so I should possibly add code chunks in here that represent the model iteration, that I don't run here, but might add it here just to show it, but without actually including it when rendering.

Also need to figure out how to do the framework with having different qmd files for investigating results from different models, and how that should be all incorporated into the thesis in the end.



```{r}
out_shrink <- readRDS("output_markov/Output_gamma_shrink.rds")
```


```{r}
print(out_shrink, probs = c(0.025, 0.975), digits_summary = 3)


```

Okay, that's a lot of parameters, but it looks mostly fine, the tau's look good. It was only 200 total iterations though. So let's take a look at the runtimes.

```{r}
library(rstan)

times <- get_elapsed_time(out_shrink)
print(times)
```

Now, let's take a look at the plot of the chains: 

```{r}
library(bayesplot)
library(ggplot2)

color_scheme_set("mix-blue-red")

chains1 <- mcmc_trace(out_shrink, pars = c("Lambda[1,1]", "Lambda[1,2]", "Lambda[3,1]", "Lambda[3,2]", "Lambda[2,3]", "Lambda[1,4]"))

chains2 <- mcmc_trace(out_shrink, pars = c("tau[1]", "tau[2]", "tau[3]", "tau[4]", "gamma_repro[3]", "gamma_surv[4]"))

# chains3 <- mcmc_trace(out_no_pos, pars= c("lp__", "h2_lv[1]", "h2_lv[2]"))

print(chains1)
print(chains2)

```


Okay, so it's not sign flipping we have, it's that the model has no idea what any of the parameters are. It's either label switching or the model just has no clue what it's doing. I think the paper might mention that it doesn't solve the rotational problem, only the dimension problem (need to verify this, because I thought the point of the paper was that we shouldn't enforce an echelon form of the loading matrix and thereby this was a way to avoid the rotational issue and hopefully solve it by not enforcing a structure, but allowing the model to select it and there hopfully being a single mode that the model lands on given the shrinkage of the variances), but I thought the shrinkage prior would also ensure we don't have label switching, but it might be an issue that the model simply isn't identifiable with this many latent variables as well. 

So, I suspect that the fix is not as simple as adding more iterations, that would also take forever. Firstly we probably need to try with Nlv = 3 max. Then, we should probably anchor the first trait to load positively on the first factor. 


Question: Maybe skip the local shrinkage and only have the global shrinkage, because all the phi values are around 1. And we haven't been doing that at all previously. 



Alright, now I will run a model with maximum 3 latent variables instead of 4, and i will not include the trait-specific (local) penalty. See if that works better. Here are the results (it also took a lot shorter, roughly 2 h 15 min for 250 iterations vs like 8 hours for 200 iterations in the previous run):

```{r}
out_no_phi <- readRDS("output_markov/Output_gamma_shrink_no_phi.rds")
```

Let's take a look at everything we can about these results then: 

```{r}
print(out_no_phi, probs = c(0.025, 0.975), digits_summary = 3)

```


```{r}
library(rstan)
library(bayesplot)
library(ggplot2)

times_no_phi <- get_elapsed_time(out_no_phi)

print(times_no_phi)

chains_no_phi1 <- mcmc_trace(out_no_phi, pars = c("Lambda[1,1]", "Lambda[1,2]", "Lambda[3,1]", "Lambda[3,2]", "Lambda[2,3]"))
chains_no_phi2 <- mcmc_trace(out_no_phi, pars = c("tau[1]", "tau[2]", "tau[3]", "gamma_repro[3]"))

print(chains_no_phi1)
print(chains_no_phi2)
```
we have sign flipping of the third latent variable i think (chain 4). now it would be very tempting to just constrain (anchor) one trait per latent variable. The issue with that is something that happened on a previous run, in which what happened was that i constrained the first trait to be positive for the first lv, and the second trait to be positive for the second latent variable. And what happened then is that it caused the body size to be the second latent variable instead of the first. 

TIP: IF I GET STUCK INSIDE A DOULBE TMUX TERMINAL JUST TYPE RESET AND ENTER INSIDE IT. 



Now I'm running it with the a_1 and a_2 pre-specified instead of estimated, also a few minor adjustments, I think. 

```{r}
out_no_phi_or_a <- readRDS("output_markov/Output_gamma_shrink_no_phi_or_a.rds")

print(out_no_phi_or_a, probs = c(0.025, 0.975), digits_summary = 3)
```

Oops, I accidentally removed the definition of the proportion of the variance explained, causing som missing values. But that can just be computed afterwards.

Are the a's necessary to estimate in order to select the number of latent variables? In any case, let's take a look at the chains for the Lambda's and the tau's as well as the gamma's.

```{r}
library(rstan)
library(bayesplot)
library(ggplot2)

times_no_phi_or_a <- get_elapsed_time(out_no_phi_or_a)
print(times_no_phi_or_a)


clean_lambda_array <- as.array(out_no_phi_or_a, 
                               pars = c("Lambda[1,1]", "Lambda[1,2]", "Lambda[1,3]", 
                                        "Lambda[2,1]", "Lambda[2,2]", "Lambda[2,3]", 
                                        "Lambda[3,1]", "Lambda[3,2]", "Lambda[3,3]"))

clean_lambda_array2 <- as.array(out_no_phi_or_a, 
                               pars = c("Lambda[4,1]", "Lambda[4,2]", "Lambda[4,3]", 
                                        "Lambda[5,1]", "Lambda[5,2]", "Lambda[5,3]"))

clean_tau_gamma_array <- as.array(out_no_phi_or_a, 
                               pars = c("tau[1]", "tau[2]", "tau[3]", "gamma_repro[1]","gamma_repro[2]","gamma_repro[3]", "gamma_surv[1]", "gamma_surv[2]", "gamma_surv[3]"))


# 2. Pass the clean array directly to mcmc_trace (no need for the 'pars' argument here)
chains_no_phi_or_a1 <- mcmc_trace(clean_lambda_array)
chains_no_phi_or_a2 <- mcmc_trace(clean_lambda_array2)
chains_no_phi_or_a3 <- mcmc_trace(clean_tau_gamma_array)

print(chains_no_phi_or_a1)
print(chains_no_phi_or_a2)
print(chains_no_phi_or_a3)

```
It is clear that the second LV has sign flipping. Otherwise the results look fine, it seems again that the tarsus is more aligned with the second latent variable than the first, which is interesting. 

```{r}

clean_h2_array <- as.array(out_no_phi_or_a, 
                               pars = c("h2_lv[1]", "h2_lv[2]", "h2_lv[3]"))

chains_no_phi_or_a4 <- mcmc_trace(clean_h2_array)

print(chains_no_phi_or_a4)


```
It's clear that the second latent variable shows sign flipping, so let's correct that and plot out the results after that change. 

```{r}
library(rstan)
library(bayesplot)

# 1. Define all the Lambda and Gamma parameters we care about
all_pars <- c(
  "Lambda[1,1]", "Lambda[1,2]", "Lambda[1,3]",
  "Lambda[2,1]", "Lambda[2,2]", "Lambda[2,3]",
  "Lambda[3,1]", "Lambda[3,2]", "Lambda[3,3]",
  "Lambda[4,1]", "Lambda[4,2]", "Lambda[4,3]",
  "Lambda[5,1]", "Lambda[5,2]", "Lambda[5,3]",
  "gamma_repro[1]", "gamma_repro[2]", "gamma_repro[3]",
  "gamma_surv[1]",  "gamma_surv[2]",  "gamma_surv[3]"
)

# Extract into a 3D array: Dimensions are [iterations, chains, parameters]
full_array <- as.array(out_no_phi_or_a, pars = all_pars)

# 2. Identify ONLY the parameters that belong to Latent Variable 2
lv2_pars <- c(
  "Lambda[1,2]", "Lambda[2,2]", "Lambda[3,2]", 
  "Lambda[4,2]", "Lambda[5,2]", 
  "gamma_repro[2]", "gamma_surv[2]"
)

# 3. Flip the signs for Chains 1 and 4
chains_to_flip <- c(2, 3)

for (p in lv2_pars) {
  # Multiply those specific chains by -1
  full_array[, chains_to_flip, p] <- full_array[, chains_to_flip, p] * -1
}

# 4. Visual Check: Plot the trace of LV2 to ensure they are now merged!
# (They should look like a single fuzzy caterpillar now)
color_scheme_set("mix-blue-red")
print(mcmc_trace(full_array, pars = c("Lambda[1,2]", "Lambda[2,2]", "Lambda[3,2]")))

# 5. Calculate the CORRECTED summary statistics
# rstan's monitor() function rebuilds the summary table (including Rhat) from our modified array
# 5. Calculate the CORRECTED summary statistics
# We explicitly set warmup = 0 so rstan knows the array is already clean
# Load the modern Stan summary package
library(posterior)

# Convert your modified array into a "draws" object
clean_draws <- as_draws_array(full_array)

# Generate a clean summary table (automatically calculates Rhat and ESS!)
corrected_summary <- summarise_draws(clean_draws)

# Print it nicely to the console
# Force it to a base R dataframe to bypass the printing bug
clean_df <- as.data.frame(corrected_summary)

# Print specific columns so it fits on your screen
print(clean_df[, c("variable", "mean", "sd", "q5", "q95", "rhat", "ess_bulk")])
```
OK, now it's looking better, although it's not converged. But it looks like the model is not strictly rejecting the third latent variable, although the shrinkage is real. 

